---
---
---
title: "Homework 3"
author: "Your Name"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
bryozoan <- read_csv("https://sta112-f21.github.io/homework/bryozoan_data.csv")
```

## Question 1

```{r}

```

Type your answer to question 1 here...

## Question 2

```{r}

```

Type your answer to question 2 here...

## Question 3

```{r}

```

Type your answer to question 3 here...

## Question 4

```{r}

```

## Question 5

```{r}

```

Type your answer to question 5 here...

## Question 6

```{r}

```

## Question 7

Type your answer to question 7 here...

## Question 8

```{r}

```

Type your answer to question 8 here...

## Question 9

Type your answer to question 9 here...

## Question 10

```{r}

```

Type your answer to question 10 here...

## Question 11

```{r}

```

Type your answer to question 11 here...

## Question 12

Type your answer to question 12 here...

## Question 13

```{r}

```

## Question 14

```{r}

```

Type your answer to question 14 here...

## Question 15

```{r}

```

Type your answer to question 15 here...

## Question 16

```{r}

```

Type your answer to question 16 here...

## Question 17

Type your answer to question 17 here...

## Question 18

* *Data matching:* 
* *Exhaustive:* 
* *Skeptical:* 
* *Second-order:* 
* *Transparency:* 
* *Reproducibility:* 

## Citations and acknowledgments

Cite any outside resources used, and acknowledge any collaborators

