---
---
---
title: "Homework 2"
author: "Your Name"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
drosophila <- read_csv("https://sta112-f21.github.io/homework/drosophila.csv")
```

## Question 1

```{r}

```

Type your answer to question 1 here...

## Question 2

```{r}

```

Type your answer to question 2 here...

## Question 3

```{r}

```

Type your answer to question 3 here...

## Question 4

```{r}

```

Type your answer to question 4 here...

## Question 5

```{r}

```

Type your answer to question 5 here...

## Question 6

```{r}

```

## Question 7

Type your answer to question 7 here...

## Question 8

```{r}

```

## Question 9

Type your answer to question 9 here...

## Question 10

* *Data matching:* 
* *Exhaustive:* 
* *Skeptical:* 
* *Second-order:* 
* *Transparency:* 10 The analysis depends entirely on a few visualizations and summary statistics, and so is very transparent.
* *Reproducibility:* 

## Citations and acknowledgments

Cite any outside resources used, and acknowledge any collaborators

