---
author: "Dr. Ciaran Evans"
title: Understanding the logistic regression model
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

## Agenda

* AWM info session on internships and REUs
    * tomorrow (Nov 4) at 6pm, Manchester 018
* Today: logistic regression interpretation

---

## Concept check

.pull-left[
Survival on the *Titanic* by passenger sex:

| | Female | Male | Total |
| --- | --- | --- | --- |
| Survived | 233 | 109 | 342 |
| Did not survive | 81 | 468 | 549 |
| Total | 314 | 577 | 891 |

]

.pull-right[
What are the odds of survival for female passengers?

.abox[
0.26
]

.bbox[
0.68
]

.cbox[
0.74
]

.dbox[
2.88
]

]

--

**Answer:** (233/314)/(81/314) = 2.88

---

## Concept check

.pull-left[
Survival on the *Titanic* by passenger sex:

| | Female | Male | Total |
| --- | --- | --- | --- |
| Survived | 233 | 109 | 342 |
| Did not survive | 81 | 468 | 549 |
| Total | 314 | 577 | 891 |

]

.pull-right[
The odds of survival for female passengers are 2.88. How would you interpret these odds?

.abox[
Female passengers are 2.88 times more likely to survive than male passengers.
]

.bbox[
Female passengers are 2.88 times more likely to survive than not survive.
]

.cbox[
The probability of survival for female passengers is 2.88.
]


]

--

**Answer:** Female passengers are 2.88 times more likely to survive than not survive.

---

## Concept check

.pull-left[
Survival on the *Titanic* by passenger sex:

| | Female | Male | Total |
| --- | --- | --- | --- |
| Survived | 233 | 109 | 342 |
| Did not survive | 81 | 468 | 549 |
| Total | 314 | 577 | 891 |

]

.pull-right[
What is the odds ratio of survival for female vs. male passengers?

.abox[
12.35
]

.bbox[
2.88
]

.cbox[
0.08
]

.dbox[
3.92
]

]

--

**Answer:** (233/81)/(109/468) = 12.35

---

## Concept check

.pull-left[
Survival on the *Titanic* by passenger sex:

| | Female | Male | Total |
| --- | --- | --- | --- |
| Survived | 233 | 109 | 342 |
| Did not survive | 81 | 468 | 549 |
| Total | 314 | 577 | 891 |

]

.pull-right[
The odds ratio of survival for female vs. male passengers is 12.35. How would you interpret this odds ratio?

.abox[
Female passengers are 12.35 times more likely to survive than male passengers.
]

.bbox[
The odds of survival are 12.35 times higher for female passengers than for male passengers.
]

.cbox[
Female passengers are 12.35 times more likely to survive than not survive.
]

.dbox[
The probability of survival for female passengers is 12.35.
]

]

--

**Answer:** The odds of survival are 12.35 times higher for female passengers than for male passengers.

---

## Concept check

.pull-left[
Now let's look at the relationship between age and survival for *Titanic* passengers. Our fitted logistic regression model is

$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = -0.057 - 0.011 \ \text{age}$,

where $\pi = P(\text{survived})$
]

.pull-right[
Based on the fitted logistic regression model, are older passengers more or less likely to survive?

.abox[
More likely
]

.cbox[
Less likely
]
]

--

**Answer:** less likely (odds of survival decreases with age, so the probability of survival decreases with age too)

---

## Concept check

.pull-left[
Now let's look at the relationship between age and survival for *Titanic* passengers. Our fitted logistic regression model is

$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = -0.057 - 0.011 \ \text{age}$,

where $\pi = P(\text{survived})$
]

.pull-right[
What are the estimated odds of survival for a newborn baby?

.abox[
-0.057
]

.bbox[
0.94
]

.cbox[
-0.011
]

.dbox[
0.99
]

]

--

**Answer:** $\exp\{-0.057\} = 0.94$

---

## Concept check

.pull-left[
Now let's look at the relationship between age and survival for *Titanic* passengers. Our fitted logistic regression model is

$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = -0.057 - 0.011 \ \text{age}$,

where $\pi = P(\text{survived})$
]

.pull-right[
What is the estimated change in the odds of survival for an increase of 1 year in age?

.abox[
The odds of surival change by a factor of -0.057
]

.bbox[
The odds of survival change by a factor of 0.94
]

.cbox[
The odds of survival change by a factor of -0.011
]

.dbox[
The odds of survival change by a factor of 0.99
]

]

--

**Answer:** $\exp\{-0.011\} = 0.99$

---

## Concept check

.pull-left[
Now let's look at the relationship between age and survival for *Titanic* passengers. Our fitted logistic regression model is

$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = -0.057 - 0.011 \ \text{age}$,

where $\pi = P(\text{survived})$
]

.pull-right[
What is the estimated *probability* of survival for a passenger 40 years old?

.abox[
0.38
]

.bbox[
0.61
]

.cbox[
-0.497
]

.dbox[
0.99
]

]

--

**Answer:** $\dfrac{\exp\{-0.057 - 0.011 \cdot 40\}}{1 + \exp\{-0.057 - 0.011 \cdot 40\}} = 0.38$

---

## Converting from log-odds to probabilities

$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = \widehat{\beta}_0 + \widehat{\beta}_1 \ x$

$\widehat{\pi} = \dfrac{\exp\{\widehat{\beta}_0 + \widehat{\beta}_1 \ x \}}{1 + \exp\{\widehat{\beta}_0 + \widehat{\beta}_1 \ x \}}$

---

## Shape of the fitted curve

How does the shape of the fitted logistic regression depend on $\widehat{\beta}_0$ and $\widehat{\beta}_1$?

.pull-left[
$\widehat{\pi} = \dfrac{\exp\{\widehat{\beta}_0 +  x \}}{1 + \exp\{\widehat{\beta}_0 + x \}} \hspace{0.5cm}$ for $\widehat{\beta}_0 = -3, -1, 1$

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
library(tidyverse)

expit <- function(p){
  return(exp(p)/(1 + exp(p)))
}

data.frame(x = seq(-5, 6, length.out=100),
           y1 = expit(-1 + seq(-5, 6, length.out=100)),
           y2 = expit(-3 + seq(-5, 6, length.out=100)),
           y3 = expit(1 + seq(-5, 6, length.out=100))) %>%
  ggplot(aes(x = x, y = y1)) +
  geom_line(lwd = 1.2) +
  geom_line(aes(y = y2), lwd=1.2, lty = 2, color="blue") +
  geom_line(aes(y = y3), lwd=1.2, lty = 3, color="red") +
  theme_bw() +
  labs(x = "x", y = "P(y = 1)") +
  annotate("text", x = -2, y = 0.4, label="1", size=8) +
  annotate("text", x = 0, y = 0.4, label="-1", size=9) +
  annotate("text", x = 3.5, y = 0.4, label="-3", size=8) +
  theme(text = element_text(size = 20))
```
]

.pull-right[
$\widehat{\pi} = \dfrac{\exp\{-1 +  \widehat{\beta}_1 \ x \}}{1 + \exp\{-1 +  \widehat{\beta}_1 \ x \}} \hspace{0.5cm}$ for $\widehat{\beta}_1 = 0.5, 1, 2$

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

data.frame(x = seq(-5, 6, length.out=100),
           y1 = expit(-1 + 0.5*seq(-5, 6, length.out=100)),
           y2 = expit(-1 + seq(-5, 6, length.out=100)),
           y3 = expit(-1 + 2*seq(-5, 6, length.out=100))) %>%
  ggplot(aes(x = x, y = y1)) +
  geom_line(lwd = 1.2) +
  geom_line(aes(y = y2), lwd=1.2, lty = 2, color="blue") +
  geom_line(aes(y = y3), lwd=1.2, lty = 3, color="red") +
  theme_bw() +
  labs(x = "x", y = "P(y = 1)") +
  annotate("text", x = 3, y = 0.75, label="0.5", size=8) +
  annotate("text", x = 1.5, y = 0.75, label="1", size=9) +
  annotate("text", x = 0.5, y = 0.75, label="2", size=8) +
  theme(text = element_text(size = 20))
```
]

---

## Fitting logistic regression in R

```{r}
library(Stat2Data)
data("MedGPA")

med_glm <- glm(Acceptance ~ GPA, data = MedGPA,
               family = binomial)
```

* `glm` stands for "Generalized Linear Model"
* Use `family = binomial` for binary responses

---

## Fitting logistic regression in R

```{r}
library(Stat2Data)
data("MedGPA")

med_glm <- glm(Acceptance ~ GPA, data = MedGPA,
               family = binomial)
med_glm
```

$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = -19.21 + 5.45 \ \text{GPA}$ 

---

## Class activity

Predicting med school acceptance from MCAT score

[https://sta112-f21.github.io/class_activities/ca_lecture_31.html](https://sta112-f21.github.io/class_activities/ca_lecture_31.html)

---

## Class activity

$\pi = P(Accepted = 1)$

$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = -8.712 + 0.246 \ \text{MCAT}$ 

What is the change in the odds of acceptance associated with an increase of 1 point on the MCAT?

--

**Answer:** An increase of 1 point on the MCAT is associated with an increase in the odds of acceptance by a factor of $\exp\{ 0.246 \} = 1.28$.

---

## Class activity

$\pi = P(Accepted = 1)$

$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = -8.712 + 0.246 \ \text{MCAT}$ 

What is the estimated probability that a student with an MCAT score of 40 is accepted?

--

**Answer:** $\widehat{\pi} = \dfrac{ \exp\{-8.712 + 0.246 \cdot 40 \} }{1 +  \exp\{-8.712 + 0.246 \cdot 40 \}} = 0.76$

So we estimate that a student with an MCAT score of 40 has a 76% chance of being accepted to medical school.

---

## Class activity

$\pi = P(Accepted = 1)$

$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = -8.712 + 0.246 \ \text{MCAT}$ 

For approximately what MCAT score would a student have a roughly 50-50 chance of being accepted to medical school?

--

**Answer:** We need odds = 1, so $\dfrac{\widehat{\pi}}{1 - \widehat{\pi}} = \exp\{ -8.712 + 0.246 \ \text{MCAT} \} = 1$

So $-8.712 + 0.246 \ \text{MCAT} = 0$, which means $\text{MCAT} = 8.712/0.246 = 35.4$