---
author: "Dr. Ciaran Evans"
title: Lab 6 Interactions and nested F-tests
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r, include=F}
library(tidyverse)
library(knitr)
library(palmerpenguins)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

fire_clusters <- read_csv("fire-clusters.csv") %>%
  mutate(proportion = BB/bld_clust)

fire_clusters <- fire_clusters %>%
  filter(proportion > 0)

fire_clusters <- fire_clusters %>%
  mutate(log_proportion = log(proportion),
         log_bld_den = log(bld_den),
         log_clust_dist = log(near_clust))

lm_dens <- lm(log_proportion ~ log_bld_den, data = fire_clusters)
```

## Agenda

* Mid-semester grades
    * Don't have lowest grade dropped
    * Don't include Project 1
* Major/minor advising and future stats courses
* Recap interactions
* Lab 6

---

## Future stats courses

Thinking about continuing with statistics? Nice! Here are some courses you could take:

* STA 247 (Design and sampling)
* STA 279 (Categorical data) 
    * Requires MST 111 (Calc 1)
* STA 310 (Probability)
    * Requires MST 112 (Calc 2)
* STA 362 (Multivariate statistics)
    * Requires MST 121 or MST 205 (Linear algebra)
* STA 363 (Statistical learning)
    * Requires MST 121 or MST 205 (Linear algebra)

---

### Exploring multivariate data: wildfire damage

**Question:** What factors are associated with building loss in US wildfires?

.footnote[
Alexandre PM, Stewart SI, Keuler NS, Clayton MK, Mockrin MH, Bar-Massada A, Syphard AD, Radeloff VC (2016). Factors related to building loss due to wildfires in the conterminous United States. Ecological Applications, 26: 2323-2338. [https://doi.org/10.1002/eap.1376](https://doi.org/10.1002/eap.1376)
]

**Data:** Area inside US wildfires from 2000 to 2010 divided into clusters of buildings. For each cluster, variables on building arrangement, vegetation, and topography are recorded. Variables include:

* *proportion*: proportion of buildings in the cluster which were burned
* *bld_den*: density of buildings in the cluster
* *bld_disp*: dispersion of buildings in the cluster
* *elev*: elevation of the cluster (meters)
* *sloped*: slope of the land (degrees)
* *CONTAG*: contagion index near the cluster
* *CONNECT*: connectance index near the cluster
* *PLAND1*: proportion of land near the cluster that is highly flammable

---

## Recap: Interaction

**Model with interaction:** slope is different for each species

$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \beta_3 \text{body mass} +$

$\hspace{2cm} \beta_4 \text{IsChinstrap} \cdot \text{body mass} + \beta_5 \text{IsGentoo} \cdot \text{body mass} + \varepsilon$

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}

penguins %>%
  drop_na() %>%
  ggplot(aes(x = body_mass_g, y = bill_length_mm, color = species)) +
  geom_point() +
  geom_smooth(method = "lm", se=F) +
  theme_bw() +
  labs(x = "Body mass (g)", y = "Bill length (mm)",
       color = "Species") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

---

## Interaction for the wildfire data


**Model:** $\ \log(\text{proportion}) = \beta_0 + \beta_1 \log(\text{building density}) + \beta_2  \log(\text{cluster distance}) +$

$\hspace{4cm} \beta_3 \log(\text{building density}) \cdot \log(\text{cluster distance}) + \varepsilon$

**Fitted model:** $\ \widehat{\log(\text{proportion})} = -2.38 - 0.93 \log(\text{building density}) + 0.086  \log(\text{cluster distance}) +$

$\hspace{2cm} 0.094 \log(\text{building density}) \cdot \log(\text{cluster distance})$

---

## Interaction for the wildfire data

**Fitted model:** $\ \widehat{\log(\text{proportion})} = -2.38 - 0.93 \log(\text{building density}) + 0.086  \log(\text{cluster distance}) +$

$\hspace{2cm} 0.094 \log(\text{building density}) \cdot \log(\text{cluster distance})$

An increase in log(building density) is associated with a greater decrease in log(proportion of buildings burned) when:

.abox[
The cluster is more isolated (greater distance to the nearest cluster)
]

.cbox[
The cluster is less isolated (smaller distance to the nearest cluster)
]

---

## Interaction for the wildfire data

**Fitted model:** $\ \widehat{\log(\text{proportion})} = -2.38 - 0.93 \log(\text{building density}) + 0.086  \log(\text{cluster distance}) +$

$\hspace{2cm} 0.094 \log(\text{building density}) \cdot \log(\text{cluster distance})$

An increase in log(building density) is associated with a greater decrease in log(proportion of buildings burned) when:

.abox[
The cluster is more isolated (greater distance to the nearest cluster)
]

.cbox[
The cluster is less isolated (smaller distance to the nearest cluster)
]

**Answer:** When the cluster is less isolated (smaller distance to the nearest cluster)

---

## In R

```{r}
lm(log_proportion ~ log_bld_den*log_clust_dist, data = fire_clusters)
```

For interaction between two variables, replace `+` with `*`


$\ \widehat{\log(\text{proportion})} = -2.38 - 0.93 \log(\text{building density}) + 0.086  \log(\text{cluster distance}) +$

$\hspace{2cm} 0.094 \log(\text{building density}) \cdot \log(\text{cluster distance})$

---

## In R

.small[
```{r, highlight.output=c(19)}
fire_lm <- lm(log_proportion ~ log_bld_den*log_clust_dist, 
              data = fire_clusters)
summary(fire_lm)
```
]

$R^2_{adj}$ has increased from about 6.3% to about 6.6%, so the interaction term explains very little additional variability.

---

### Lab 6: Diversity, Testosterone, and Group Performance

**Question:** Does the relationship between group diversity and group performance at a task depend on the levels of testosterone in the group?

**Data:** 74 groups of MBA students completing a week-long project on maximizing financial performance for a blood-testing lab. Each row in the data is a group, with the variables

* *performance*: the final performance of the group in the project
* *size*: the number of group members (between 3 and 6)
* *females*: the number of female group members
* *diversity*: a measure of group diversity that summarizes information on the gender, ethnicity, and country of origin for the group members
* *testosterone*: the average testosterone levels (in pg/mL) for the group members

--

.question[
What regression model could we use to explore whether the relationship between diversity and performance is different for different levels of testosterone? Should we include *size* and *females* in the model?
]

---

### Lab 6: Diversity, Testosterone, and Group Performance

We will use a model like 

$\text{performance} = \beta_0 + \beta_1 \text{size} + \beta_2 \text{females} + \beta_3 \text{diversity} + \beta_4 \text{testosterone}$

$\hspace{4cm} + \beta_5 \text{diversity} \cdot \text{testosterone} + \varepsilon$

**Lab 6:** Practice with

* Fitting, interpreting, and assessing multiple regression models
* Including interaction terms
* Nested F-tests