---
author: "Dr. Ciaran Evans"
title: Interpretation and Lab 3
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

## Agenda

* HW 3 released today; due in **two** weeks
    * longer assignment than normal
    * start early
    * we will cover some of the material next week
    * really cool example where linear regression is used in a real research paper
* Interpreting linear regression fit
* Lab 3

---

## Kent sparrows data

```{r echo=F, message=F, fig.align='center', fig.width = 5, fig.height=4}
library(Stat2Data)
library(tidyverse)
data(Sparrows)

Sparrows %>%
  ggplot(aes(x = WingLength, y = Weight)) +
  geom_point() +
  geom_smooth(method = "lm", se=F) +
  labs(y = "Weight (g)", 
       x = "Wing Length (cm)") +
  theme_bw() +
  theme(text = element_text(size=20))
```

```{r}
lm(Weight ~ WingLength, data = Sparrows)
```

---

## Kent sparrows data

```{r echo=F, message=F, fig.align='center', fig.width = 5, fig.height=4}
Sparrows %>%
  ggplot(aes(x = WingLength, y = Weight)) +
  geom_point() +
  geom_smooth(method = "lm", se=F) +
  labs(y = "Weight (g)", 
       x = "Wing Length (cm)") +
  theme_bw() +
  theme(text = element_text(size=20))
```

$\widehat{\text{weight}} = 1.366 + 0.467(\text{wing length})$

---

## $\widehat{\beta}_0$

```{r echo=F, message=F, fig.align='center', fig.width = 5, fig.height=4}
Sparrows %>%
  ggplot(aes(x = WingLength, y = Weight)) +
  geom_point() +
  geom_smooth(method = "lm", se=F) +
  labs(y = "Weight (g)", 
       x = "Wing Length (cm)") +
  theme_bw() +
  theme(text = element_text(size=20))
```

$\widehat{\text{weight}} = 1.366 + 0.467(\text{wing length})$

$\widehat{\beta}_0 = 1.366$

.question[
The estimated weight for a sparrow with wing length 0cm is 1.366g
]

---

## $\widehat{\beta}_0$

```{r echo=F, message=F, fig.align='center', fig.width = 5, fig.height=4}
Sparrows %>%
  ggplot(aes(x = WingLength, y = Weight)) +
  geom_point() +
  geom_smooth(method = "lm", se=F) +
  labs(y = "Weight (g)", 
       x = "Wing Length (cm)") +
  theme_bw() +
  theme(text = element_text(size=20))
```

$\widehat{\beta}_0 = 1.366$

The estimated weight for a sparrow with wing length 0cm is 1.366g

.question[
If $x = 0$ is outside the range of the data, then $\widehat{\beta}_0$ might not have a very meaningful interpretation.
]

---

## $\widehat{\beta}_1$

```{r echo=F, message=F, fig.align='center', fig.width = 5, fig.height=4}
Sparrows %>%
  ggplot(aes(x = WingLength, y = Weight)) +
  geom_point() +
  geom_smooth(method = "lm", se=F) +
  labs(y = "Weight (g)", 
       x = "Wing Length (cm)") +
  theme_bw() +
  theme(text = element_text(size=20))
```

$\widehat{\text{weight}} = 1.366 + 0.467(\text{wing length})$

$\widehat{\beta}_1 = 0.467$

.question[
An increase of one cm in wing length is associated with an increase in 0.467g, on average, in weight.
]

---

## Be careful with causality!

$\widehat{\text{weight}} = 1.366 + 0.467(\text{wing length})$

**Some appropriate interpretations:**

* An increase of one cm in wing length is associated with an increase in 0.467g, on average, in weight.
* If sparrow A has a wing length that is one cm longer than sparrow B's wing length, we predict that sparrow A weighs 0.467g more than sparrow B.

--

**Some inappropriate interpretations:** (except in special cases)

* If a sparrow's wings grow by 1cm, then its weight will increase by 0.467g.
* An increase of 1cm in wing length causes an increase of 0.467g in weight.

--

**When are causal interpretations justified?**

* In randomized experiments
* In some observational studies, if we appropriately account for confounding variables

---

## Lab 3

Practice with linear regression on the penguins data.

[https://sta112-f21.github.io/labs/lab_3.html](https://sta112-f21.github.io/labs/lab_3.html)