---
author: "Dr. Ciaran Evans"
title: Polynomial regression and exploring multivariate data
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

## Agenda

* Project 1 comments:
    * It is ok if your conclusion is that there is no relationship (provided you provide sound statistical justification for your conclusion, as usual)
    * $\log(0) = -\infty$. Some options:
        * Add a small number before transforming, e.g. $\log(variable + 1)$ instead of $\log(variable)$
        * Try a different transformation
        * Remove rows where $variable = 0$. Not necessarily a good idea, needs to be justified.
    * Only simple linear regression is required. You are welcome to fit a multiple regression model if you like, but it is not required
* Today:
    * Go over Quiz 3
    * Polynomial regression
    * Exploring multiple regression with wildfire data

---

## Quiz 3

Errors:

* Interpretation of the p-value
    * p-value is the probability of "our data or more extreme" if $H_0$ is true; not the probability that $H_0$ is true
* Interpretation of the intercept/slope
* Assessment of normality
    * Assumption is that the residuals are normal, not the predictors
    * (Also better to check with QQ plot, not histogram)
* Interpretation of the boxplots
    * Doesn't make sense to think about shape (e.g., linearity) for an un-ordered categorical predictor
* Conclusion
    * Collaborator is implying that there is a causal relationship between sleep, homework location, and GPA

---

## German carbon dioxide levels

German carbon dioxide levels between April and November 2001:

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
library(Stat2Data)
library(tidyverse)
data("CO2Germany")
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point() +
  labs(x = "Day (since start of the year)",
       y = "Carbon dioxide (ppm) in atmosphere") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

--

.question[
The relationship looks like it could be quadratic. How can we fit that?
]

---

## Quadratic regression model

**Model:** $\text{CO}_2 = \beta_0 + \beta_1 \text{Day} + \beta_2 \text{Day}^2 + \varepsilon$


```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
quad_lm <- lm(CO2 ~ Day + I(Day^2), data = CO2Germany)

CO2Germany %>%
  mutate(pred = predict(quad_lm)) %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point() +
  geom_line(aes(y = pred), color = "blue", lwd=1.5) +
  labs(x = "Day (since start of the year)",
       y = "Carbon dioxide (ppm) in atmosphere") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

**Fitted model:** $\widehat{\text{CO}}_2 = 414.975 - 0.476 \ \text{Day} + 0.001 \ \text{Day}^2$

---

## Assumptions

We make the usual assumptions:

* Shape (in this case, the shape assumption is that our polynomial is right)
* Constant variance
* Independence
* Randomness
* Zero mean
* Normality

---

## Checking assumptions

.pull-left[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
CO2Germany %>%
  mutate(pred = predict(quad_lm),
         res = resid(quad_lm)) %>%
  ggplot(aes(x = pred,
             y = res)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(x = "Predicted carbon dioxide level (ppm)",
       y = "Residuals") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

.pull-right[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
CO2Germany %>%
  mutate(res = resid(quad_lm)) %>%
  ggplot(aes(sample = res)) +
  geom_qq() +
  geom_qq_line() +
  labs(x = "Theoretical normal quantiles",
       y = "Observed residual quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

---

## In R

```{r}
co2_lm <- lm(CO2 ~ Day + I(Day^2), data = CO2Germany)
summary(co2_lm)
```

---

## In R

```{r, eval=F}
co2_lm <- lm(CO2 ~ Day + I(Day^2), data = CO2Germany)
summary(co2_lm)
```

`I(...)` is used in formulas in R when we want to evaluate the expression arithmetically

---

## Higher-order polynomial?


**Model:** $\text{CO}_2 = \beta_0 + \beta_1 \text{Day} + \beta_2 \text{Day}^2 + \beta_3 \text{Day}^3 + \varepsilon$


```{r echo=F, message=F, warning=F, fig.align='center', fig.width=8, fig.height=5}
cub_lm <- lm(CO2 ~ Day + I(Day^2) + I(Day^3), data = CO2Germany)

CO2Germany %>%
  mutate(pred = predict(cub_lm),
         type = "Cubic model") %>%
  rbind(CO2Germany %>%
          mutate(pred = predict(quad_lm),
                 type = "Quadratic model")) %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point() +
  geom_line(aes(y = pred, color = type), lwd=1.5) +
  labs(x = "Day (since start of the year)",
       y = "Carbon dioxide (ppm) in atmosphere",
       color = "") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

**Fitted model:** $\widehat{\text{CO}}_2 = 406.7 - 0.340 \ \text{Day} + 0.00047 \ \text{Day}^2 + 0.0000011 \ \text{Day}^3$

A cubic model doesn't look necessary.

---

## In R

```{r}
co2_lm_cubic <- lm(CO2 ~ Day + I(Day^2) + I(Day^3), data = CO2Germany)
summary(co2_lm_cubic)
```

---

### Exploring multivariate data: wildfire damage

**Question:** What factors are associated with building loss in US wildfires?

.footnote[
Alexandre PM, Stewart SI, Keuler NS, Clayton MK, Mockrin MH, Bar-Massada A, Syphard AD, Radeloff VC (2016). Factors related to building loss due to wildfires in the conterminous United States. Ecological Applications, 26: 2323-2338. [https://doi.org/10.1002/eap.1376](https://doi.org/10.1002/eap.1376)
]

**Data:** Area inside US wildfires from 2000 to 2010 divided into clusters of buildings. For each cluster, variables on building arrangement, vegetation, and topography are recorded. Variables include:

* *proportion*: proportion of buildings in the cluster which were burned
* *bld_den*: density of buildings in the cluster
* *bld_disp*: dispersion of buildings in the cluster
* *elev*: elevation of the cluster (meters)
* *sloped*: slope of the land (degrees)
* *CONTAG*: contagion index near the cluster
* *CONNECT*: connectance index near the cluster
* *PLAND1*: proportion of land near the cluster that is highly flammable

---

### Exploring multivariate data: wildfire damage

**Question:** What factors are associated with building loss in US wildfires?

**Rough plan:** We have a lot of variables to choose from, and no specific model in mind.

* Look at the proportion of buildings burned in each cluster
* Look at the relationships between *proportion* and the other variables
* Pick a variable that is interesting and makes sense, and start building a model to predict *proportion*

---

## Proportion of buildings burned

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
fire_clusters <- read_csv("fire-clusters.csv") %>%
  mutate(proportion = BB/bld_clust)

fire_clusters %>%
  ggplot(aes(x = proportion)) +
  geom_histogram() +
  labs(x = "Proportion of buildings burned") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

.question[
What do you notice about the distribution? What else would you want to know about this variable?
]

--

* Highly right-skewed
* How many clusters have 0 buildings burned?

---

## Proportion of buildings burned

```{r}
fire_clusters %>% 
  count()
fire_clusters %>%
  filter(proportion == 0) %>%
  count()
```

.question[
So many 0s could be an issue -- it messes with our linear model fit
]

---

## Revising the plan

**Revised plan:** Build two models:

* One model to predict whether any buildings were burned
* One model to predict the proportion of buildings burned, for those clusters where at least one building burned

---

## Revising the plan

**Revised plan:** Build two models:

* One model to predict whether any buildings were burned (logistic regression -- later this semester)
* One model to predict the proportion of buildings burned, for those clusters where at least one building burned (multiple linear regression with a subset of the data)

---

## Proportion of buildings burned

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
fire_clusters <- fire_clusters %>%
  filter(proportion > 0)

fire_clusters %>%
  ggplot(aes(x = proportion)) +
  geom_histogram() +
  labs(x = "Proportion of buildings burned, \n for clusters with at least one building burned") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

---

## Looking at relationships: pairs plot

.question[
Where do we start?
]

---

## Looking at relationships: pairs plot

A **pairs plot** lets us visualize pairwise relationships in the data:

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=9, fig.height=6}
library(GGally)

fire_clusters %>%
  select(proportion, bld_den, bld_disp,
         elev, sloped, CONTAG, CONNECT, PLAND1) %>%
  ggpairs(lower = list(continuous = "cor"),
          upper = list(continuous = "points")) +
  theme_minimal()
```

---

## Simple linear regression model

$\text{proportion} = \beta_0 + \beta_1 \text{building density} + \varepsilon$

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}

fire_clusters %>%
  ggplot(aes(x = bld_den, y = proportion)) +
  geom_point() +
  geom_smooth(method="lm", se=F) +
  labs(x = "Building density in cluster",
       y = "Proportion of buildings burned") +
  theme_bw() +
  theme(text = element_text(size = 20))

lm_dens <- lm(proportion ~ bld_den, data = fire_clusters)
```

$\widehat{\text{proportion}} = 0.229 - 0.003 \ \text{building density}$

---

## Diagnostics

.pull-left[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
fire_clusters %>%
  mutate(pred = predict(lm_dens),
         res = resid(lm_dens)) %>%
  ggplot(aes(x = pred,
             y = res)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(x = "Predicted proportion of buildings burned",
       y = "Residuals") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

.pull-right[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
fire_clusters %>%
  mutate(res = resid(lm_dens)) %>%
  ggplot(aes(sample = res)) +
  geom_qq() +
  geom_qq_line() +
  labs(x = "Theoretical normal quantiles",
       y = "Observed residual quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

.question[
What can I try to address these issues?
]

---

## Log transformations

$\log(\text{proportion}) = \beta_0 + \beta_1 \log(\text{building density}) + \varepsilon$

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}

fire_clusters %>%
  ggplot(aes(x = log(bld_den), y = log(proportion))) +
  geom_point() +
  geom_smooth(method="lm", se=F) +
  labs(x = "log(Building density in cluster)",
       y = "log(Proportion of buildings burned)") +
  theme_bw() +
  theme(text = element_text(size = 20))

fire_clusters <- fire_clusters %>%
  mutate(log_proportion = log(proportion),
         log_bld_den = log(bld_den))

lm_dens <- lm(log_proportion ~ log_bld_den, data = fire_clusters)
```

$\widehat{\log(\text{proportion})} = -2.04 - 0.524 \ \log(\text{building density})$

---

## Diagnostics

.pull-left[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
fire_clusters %>%
  mutate(pred = predict(lm_dens),
         res = resid(lm_dens)) %>%
  ggplot(aes(x = pred,
             y = res)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(x = "Predicted log proportion",
       y = "Residuals") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

.pull-right[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
fire_clusters %>%
  mutate(res = resid(lm_dens)) %>%
  ggplot(aes(sample = res)) +
  geom_qq() +
  geom_qq_line() +
  labs(x = "Theoretical normal quantiles",
       y = "Observed residual quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

---

## How useful is the model?

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}

fire_clusters %>%
  ggplot(aes(x = log(bld_den), y = log(proportion))) +
  geom_point() +
  geom_smooth(method="lm", se=F) +
  labs(x = "log(Building density in cluster)",
       y = "log(Proportion of buildings burned)") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

---

## How useful is the model?

**Coefficient of determination** ( $R^2$ ): the proportion of total variability in the response variable explained by the regression model.

$R^2 = \dfrac{\text{Variability explained by model}}{\text{Total variability in y}} = \dfrac{\sum \limits_{i=1}^n (\widehat{y}_i - \overline{y})^2}{\sum \limits_{i=1}^n (y_i - \overline{y})^2} = \dfrac{SSModel}{SSTotal}$

**Note:** For simple linear regression, $R^2 = r^2$ (the squared correlation coefficient)

---

### How useful is the model?

.small[
```{r, highlight.output=c(17)}
lm_dens <- lm(log_proportion ~ log_bld_den, data = fire_clusters)
summary(lm_dens)
```
]

The p-value for the slope might be low, but log(density) only explains 5% of the variability in log(proportion).

---

## Interpretation

$\widehat{\log(\text{proportion})} = -2.04 - 0.524 \ \log(\text{building density})$

.question[
Clusters with a higher density of buildings tend to have a lower proportion of buildings burned. Does this match what you expect?
]

---

## Simpson's paradox

An example with the penguin data:

.pull-left[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
library(palmerpenguins)

penguins %>%
  drop_na() %>%
  ggplot(aes(x = bill_depth_mm, y = bill_length_mm)) +
  geom_point() +
  geom_smooth(method="lm", se=F) +
  labs(x = "Bill depth (mm)",
       y = "Bill length (mm)") +
  theme_bw() +
  theme(text = element_text(size = 20))

```
]


.pull-right[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
library(palmerpenguins)

penguins %>%
  drop_na() %>%
  ggplot(aes(x = bill_depth_mm, y = bill_length_mm,
             color = species)) +
  geom_point() +
  geom_smooth(method="lm", se=F) +
  labs(x = "Bill depth (mm)",
       y = "Bill length (mm)",
       color = "Species") +
  theme_bw() +
  theme(text = element_text(size = 20))

```
]

.question[
The relationship between two variables can depend on other variables. Let's add another variable to the model. 
]

---

## Adding building dispersion

$\log(\text{proportion}) = \beta_0 + \beta_1 \log(\text{building density}) + \beta_2 \ \text{building dispersion} + \varepsilon$

.small[
```{r, highlight.output=c(18)}
lm_dens_disp <- lm(log_proportion ~ log_bld_den + bld_disp, data = fire_clusters)
summary(lm_dens_disp)
```
]

--

$R^2$ has increased slightly (from 0.05 to 0.065). We explain *slightly* more variability by adding building dispersion.

---

## $R^2_{adj}$

$R^2 = \dfrac{\sum \limits_{i=1}^n (\widehat{y}_i - \overline{y})^2}{\sum \limits_{i=1}^n (y_i - \overline{y})^2} = \dfrac{SSModel}{SSTotal} = 1 - \dfrac{SSE}{SSTotal}$

.question[
$R^2$ will **never** decrease when we add more predictors to the model.
]

--

**Adjusted $R^2$ **: $R^2_{adj} = 1 - \dfrac{SSE/(n - p)}{SSTotal/(n-1)} = 1 - \dfrac{\widehat{\sigma}^2_{\varepsilon}}{\widehat{\text{Var}}(y)}$

---


## $R^2_{adj}$

.small[
```{r, output.lines=10:18, highlight.output=c(10)}
lm_dens_disp <- lm(log_proportion ~ log_bld_den + bld_disp, data = fire_clusters)
summary(lm_dens_disp)
```
]

$R^2 = 0.065$, $R^2_{adj} = 0.061$

* $R^2_{adj} < R^2$ (always)
* $R^2_{adj}$ accounts for the number of parameters in the model
* $R^2$ *never* decreases when predictors are added
* $R^2_{adj}$ *can decrease* when unhelpful predictors are added to the model


---

## $R^2_{adj}$

.small[
```{r, output.lines=17:18, highlight.output=c(3)}
lm_dens_disp <- lm(log_proportion ~ log_bld_den + bld_disp, data = fire_clusters)
summary(lm_dens_disp)
```
]

.small[
```{r, output.lines=16:17, highlight.output=c(3)}
lm_dens <- lm(log_proportion ~ log_bld_den, data = fire_clusters)
summary(lm_dens)
```
]

Adding building dispersion increases $R^2_{adj}$ very slightly, from 0.049 to 0.061. But the $R^2_{adj}$ is still small.

---

## Interpretation

$\widehat{\log(\text{proportion})} = -0.85 - 0.42 \log(\text{building density}) - 2.06 \ \text{building dispersion}$

* Holding building dispersion fixed, an increase of one unit in $\log(\text{building density})$ is associated with a decrease of 0.42 units in $\log(\text{proportion})$, on average
* Holding building density constant, an increase of one unit in building dispersion is associated with a decrease of 2.06 units in $\log(\text{proportion})$, on average

.question[
Does this match what you expect?
]

---

## Back to the model

$\log(\text{proportion}) = \beta_0 + \beta_1 \log(\text{building density}) + \beta_2 \ \text{building dispersion} + \varepsilon$

.question[
Our model assumes that the relationship between building density and the proportion of buildings burned is the same regardless of dispersion. Is this what we should expect?
]

--

Next time: interaction



