---
author: "Dr. Ciaran Evans"
title: Model Assumptions and Diagnostics
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

## Agenda

* Course feedback:
    * adding textbook references to course webpage
    * class activities vs. labs vs. homeworks
        * class activities: learn fundamental skills
        * labs: practice combining fundamental skills in context
        * homeworks: apply skills to real data and real questions
* TA: Johe Johnson
    * gives feedback on homework in comments
    * MSC hours: Monday 12-2pm, Thursday 3-4pm
* Quiz 2 next Wednesday (September 22)
    * Covers through today
* Today:
    * Recap intro to linear regression
    * Assumptions and diagnostics


---

## Recap: regression

**Regression:** model the relationship between predictor $x$ and response $y$

**Regression model:** $y = f(x) + \varepsilon$

* $f$ = systematic information that $x$ provides about $y$
* $\varepsilon$ = random noise

--

**Linear regression:** $y = \beta_0 + \beta_1 x + \varepsilon$

---

## Recap: linear regression

**Observed data (sample):**

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
library(Stat2Data)
library(tidyverse)
data(Sparrows)

Sparrows %>%
  ggplot(aes(x = WingLength, y = Weight)) +
  geom_point() +
  labs(y = "Weight (g)", 
       x = "Wing Length (cm)") +
  theme_bw() +
  theme(text = element_text(size=20))
```

--

**Regression model (population):**

.center[
$\text{Weight} = \beta_0 + \beta_1 \text{WingLength} + \varepsilon$
]

--

**Estimated regression line (from sample):**

.center[
$\widehat{\text{Weight}} = \widehat{\beta}_0 + \widehat{\beta}_1 \text{WingLength}$
]

---

## Recap: fitting linear regression

**Model:**
.center[
$\text{Weight} = \beta_0 + \beta_1 \text{WingLength} + \varepsilon$
]

**Fitting the model:**

```{r}
lm(Weight ~ WingLength, data = Sparrows)
```

--

**Estimated regression line:**

.center[
$\widehat{\text{Weight}} = 1.366 + 0.467 \ \text{WingLength}$
]

---

## Recap: interpretation

.center[
$\widehat{\text{Weight}} = 1.366 + 0.467 \ \text{WingLength}$
]

**Interpreting $\widehat{\beta}_0$:** If wing length = 0cm, we predict a weight of 1.366 g.

**Interpreting $\widehat{\beta}_1$:** An increase of 1cm in wing length is associated with an increase of 0.467 g, on average, in weight

---

## Concept check

Interested in the relationship between mileage and price for used Honda Accords.

.pull-left[
**Observed data (sample):**

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
data("AccordPrice")
AccordPrice %>%
  ggplot(aes(x = Mileage, y = Price)) +
  geom_point() +
  labs(x = "Number of miles (in 1000's)",
       y = "Price (in $1000's)") +
  theme_bw() +
  theme(text = element_text(size=20))
```
]


.pull-right[
We will use linear regression. Write the regression model:

.abox[
$\text{price} = \beta_0 + \beta_1 \text{miles}$
]

.bbox[
$\text{price} = \beta_0 + \beta_1 \text{miles} + \varepsilon$
]

.cbox[
$\widehat{\text{price}} = \widehat{\beta}_0 + \widehat{\beta}_1 \text{miles}$
]

.dbox[
$\widehat{\text{price}} = \widehat{\beta}_0 + \widehat{\beta}_1 \text{miles} + \varepsilon$
]

]

---

## Concept check

Interested in the relationship between mileage and price for used Honda Accords.

.pull-left[
**Observed data (sample):**

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
data("AccordPrice")
AccordPrice %>%
  ggplot(aes(x = Mileage, y = Price)) +
  geom_point() +
  labs(x = "Number of miles (in 1000's)",
       y = "Price (in $1000's)") +
  theme_bw() +
  theme(text = element_text(size=20))
```
]


.pull-right[
We will use linear regression. Write the regression model:

.abox[
$\text{price} = \beta_0 + \beta_1 \text{miles}$
]

.bbox[
$\text{price} = \beta_0 + \beta_1 \text{miles} + \varepsilon$
]

.cbox[
$\widehat{\text{price}} = \widehat{\beta}_0 + \widehat{\beta}_1 \text{miles}$
]

.dbox[
$\widehat{\text{price}} = \widehat{\beta}_0 + \widehat{\beta}_1 \text{miles} + \varepsilon$
]

]

Solution: $\text{price} = \beta_0 + \beta_1 \text{miles} + \varepsilon$ 

Remember that the regression *model* describes our model for the relationship in the population.

---

## Concept check

.pull-left[
**Observed data (sample):**

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
AccordPrice %>%
  ggplot(aes(x = Mileage, y = Price)) +
  geom_point() +
  labs(x = "Number of miles (in 1000's)",
       y = "Price (in $1000's)") +
  theme_bw() +
  theme(text = element_text(size=20))
```

Least-squares fit:

```{r, echo=F}
lm(Price ~ Mileage, data = AccordPrice)
```
]


.pull-right[
What is the equation of the fitted regression line?

.abox[
$\widehat{\text{price}} = 20.81 - 0.12 \ \text{miles}$
]

.bbox[
$\text{price} = \beta_0 + \beta_1 \text{miles}$
]

.cbox[
$\widehat{\text{price}} = -0.12 + 20.81 \ \text{miles}$
]

.dbox[
$\widehat{\text{price}} = 20.81 - 0.12 \ \widehat{\text{miles}}$
]

]

---

## Concept check

.pull-left[
**Observed data (sample):**

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
AccordPrice %>%
  ggplot(aes(x = Mileage, y = Price)) +
  geom_point() +
  labs(x = "Number of miles (in 1000's)",
       y = "Price (in $1000's)") +
  theme_bw() +
  theme(text = element_text(size=20))
```

]


.pull-right[
What is the equation of the fitted regression line?

.abox[
$\widehat{\text{price}} = 20.81 - 0.12 \ \text{miles}$
]

.bbox[
$\text{price} = \beta_0 + \beta_1 \text{miles}$
]

.cbox[
$\widehat{\text{price}} = -0.12 + 20.81 \ \text{miles}$
]

.dbox[
$\widehat{\text{price}} = 20.81 - 0.12 \ \widehat{\text{miles}}$
]

]

Solution: $\widehat{\text{price}} = 20.81 - 0.12 \ \text{miles}$

---

## Concept check

.pull-left[
**Fitted line:**

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F}
AccordPrice %>%
  ggplot(aes(x = Mileage, y = Price)) +
  geom_point() +
  geom_smooth(se=F, method="lm") +
  labs(x = "Number of miles (in 1000's)",
       y = "Price (in $1000's)") +
  theme_bw() +
  theme(text = element_text(size=20))
```

$\widehat{\text{price}} = 20.81 - 0.12 \ \text{miles}$

]


.pull-right[
Interpret the estimated <ins>intercept</ins>:

.abox[
The estimated mileage for a car which costs $0 is 20,810 miles
]

.bbox[
The intercept cannot be interpreted in this situation
]

.cbox[
The estimated price for a car with 0 miles is $20,810
]

.dbox[
Every additional 1000 miles on the car is associated with a decrease of $120 in price
]

]

---

## Concept check

.pull-left[
**Fitted line:**

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F}
AccordPrice %>%
  ggplot(aes(x = Mileage, y = Price)) +
  geom_point() +
  geom_smooth(se=F, method="lm") +
  labs(x = "Number of miles (in 1000's)",
       y = "Price (in $1000's)") +
  theme_bw() +
  theme(text = element_text(size=20))
```

$\widehat{\text{price}} = 20.81 - 0.12 \ \text{miles}$

]


.pull-right[
Interpret the estimated <ins>intercept</ins>:

.abox[
The estimated mileage for a car which costs $0 is 20,810 miles
]

.bbox[
The intercept cannot be interpreted in this situation
]

.cbox[
The estimated price for a car with 0 miles is $20,810
]

.dbox[
Every additional 1000 miles on the car is associated with a decrease of $120 in price
]

]

Solution: The estimated price for a car with 0 miles is $20,810

---

## Concept check

.pull-left[
**Fitted line:**

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F}
AccordPrice %>%
  ggplot(aes(x = Mileage, y = Price)) +
  geom_point() +
  geom_smooth(se=F, method="lm") +
  labs(x = "Number of miles (in 1000's)",
       y = "Price (in $1000's)") +
  theme_bw() +
  theme(text = element_text(size=20))
```

$\widehat{\text{price}} = 20.81 - 0.12 \ \text{miles}$

]


.pull-right[
Interpret the estimated <ins>slope</ins>:

.abox[
Every additional 1000 miles on the car is associated with an increase of $120 in price, on average
]

.bbox[
Driving another 1000 miles causes the car to increase $120 in price
]

.cbox[
Driving another 1000 miles causes the car to decrease $120 in price
]

.dbox[
Every additional 1000 miles on the car is associated with a decrease of $120 in price, on average
]

]

---

## Concept check

.pull-left[
**Fitted line:**

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F}
AccordPrice %>%
  ggplot(aes(x = Mileage, y = Price)) +
  geom_point() +
  geom_smooth(se=F, method="lm") +
  labs(x = "Number of miles (in 1000's)",
       y = "Price (in $1000's)") +
  theme_bw() +
  theme(text = element_text(size=20))
```

$\widehat{\text{price}} = 20.81 - 0.12 \ \text{miles}$

]


.pull-right[
Interpret the estimated <ins>slope</ins>:

.abox[
Every additional 1000 miles on the car is associated with an increase of $120 in price, on average
]

.bbox[
Driving another 1000 miles causes the car to increase $120 in price
]

.cbox[
Driving another 1000 miles causes the car to decrease $120 in price
]

.dbox[
Every additional 1000 miles on the car is associated with a decrease of $120 in price, on average
]

]

Solution: Every additional 1000 miles on the car is associated with a decrease of $120 in price, on average

---

## Uses for regression

**Prediction:** How much should I expect to pay for a used Honda Accord with 65,000 miles?

--

**Inference:** Is the number of miles on a used Honda Accord negatively associated with sales price?

--

**Causal inference:** If I drive my used Honda Accord across the country, how will that change its value?

--

Now suppose I fit a linear regression regression model to the Accord prices data, using least squares. What assumptions do I need for each use of regression to be valid?

---

## Formal assumptions for inference

* Shape
* Constant variance
* Independence
* Randomness
* Zero mean
* Normality

---

## Formal assumptions for inference

* **Shape**
* **Constant variance**
* Independence
* Randomness
* Zero mean
* Normality

---

## Shape

**Assumption:** the shape of the regression model is at least approximately correct

--

Linear regression model:

.center[
$y = \beta_0 + \beta_1 x + \varepsilon$
]

When using linear regression, the shape assumption is that the relationship is (at least approximately) linear.

.pull-left[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  labs(title = "Relationship could be linear") +
  theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=20))
```
]

.pull-right[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
data.frame(x = 1:100,
           y = log(10*(1:100)) + rnorm(100, sd=0.3)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  labs(title = "Relationship probably not linear") +
  theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=20))
```
]

---

## Constant variance

.center[
$y = \beta_0 + \beta_1 x + \varepsilon$
]

**Assumption:** Variance of the noise $\varepsilon$ is the same for all values of the predictor $x$

.pull-left[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  labs(title = "Noise variance looks constant") +
  theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=20))
```
]

.pull-right[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20)*(1:100)*0.02) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  labs(title = "Noise variance does not look constant") +
  theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=20))
```
]

---

## Assessing shape and constant variance: residual plots

**Residual plot:** plot residuals $y - \widehat{y}$ on vertical axis, and predicted values $\widehat{y}$ on horizontal axis. Add a horizontal line at 0.

```{r, echo = F, message=F, fig.align='center', fig.width=6, fig.height=4}
library(latex2exp)
data.frame(x = 1:100,
           y = rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(title = "Residual plot",
       x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$")) +
  theme_bw() +
  theme(text = element_text(size=20))
```

---

## Residual plots

```{r, echo = F, message=F, fig.align='center', fig.width=6, fig.height=4}
data.frame(x = 1:100,
           y = rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(title = "Residual plot",
       x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$")) +
  theme_bw() +
  theme(text = element_text(size=20))
```

Shape assumption is reasonable if: 
* residuals appear to be scattered randomly above and below 0
* no clear patterns in the residuals

Constant variance assumption is reasonable if:
* the band of residuals has relatively constant width

---

## Concept check

.pull-left[
```{r, echo = F, message=F, fig.align='center', fig.width=6, fig.height=6}
data.frame(x = 1:100,
           y = rnorm(100, sd=20)*(1:100)*0.02) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=1.5) +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(title = "Residual plot",
       x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$")) +
  theme_bw() +
  theme(text = element_text(size=20))
```
]

.pull-right[
Which assumption looks reasonable?

.abox[
Shape
]

.bbox[
Constant variance
]

.cbox[
Both shape and constant variance
]

.dbox[
Neither shape nor constant variance
]

]

---

## Concept check

.pull-left[
```{r, echo = F, message=F, fig.align='center', fig.width=6, fig.height=6}
data.frame(x = 1:100,
           y = rnorm(100, sd=20)*(1:100)*0.02) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=1.5) +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(title = "Residual plot",
       x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$")) +
  theme_bw() +
  theme(text = element_text(size=20))
```
]

.pull-right[
Which assumption looks reasonable?

.abox[
Shape
]

.bbox[
Constant variance
]

.cbox[
Both shape and constant variance
]

.dbox[
Neither shape nor constant variance
]

]

Solution: the constant variance assumption is violated, because the spread of the residuals changes as we move along the x-axis. However, the shape assumption is reasonable, because the residuals are randomly scattered above and below 0 with no clear pattern.

---

## Concept check

.pull-left[
```{r, echo = F, message=F, fig.align='center', fig.width=6, fig.height=6}
sim_dat <- data.frame(x = seq(-50, 50),
           y = 0.02*seq(-50, 50)^2 + rnorm(101, sd=10)*(1:101)*0.02)
sim_reg <- lm(y ~ x, data = sim_dat)

data.frame(pred = sim_reg$fitted.values,
           res = sim_reg$residuals) %>%
  ggplot(aes(x = pred, y = res)) +
  geom_point(size=1.5) +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(title = "Residual plot",
       x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$")) +
  theme_bw() +
  theme(text = element_text(size=20))
```
]

.pull-right[
Which assumption looks reasonable?

.abox[
Shape
]

.bbox[
Constant variance
]

.cbox[
Both shape and constant variance
]

.dbox[
Neither shape nor constant variance
]

]

---

## Concept check

.pull-left[
```{r, echo = F, message=F, fig.align='center', fig.width=6, fig.height=6}
sim_dat <- data.frame(x = seq(-50, 50),
           y = 0.02*seq(-50, 50)^2 + rnorm(101, sd=10)*(1:101)*0.02)
sim_reg <- lm(y ~ x, data = sim_dat)

data.frame(pred = sim_reg$fitted.values,
           res = sim_reg$residuals) %>%
  ggplot(aes(x = pred, y = res)) +
  geom_point(size=1.5) +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(title = "Residual plot",
       x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$")) +
  theme_bw() +
  theme(text = element_text(size=20))
```
]

.pull-right[
Which assumption looks reasonable?

.abox[
Shape
]

.bbox[
Constant variance
]

.cbox[
Both shape and constant variance
]

.dbox[
Neither shape nor constant variance
]

]

Solution: neither the shape nor constant variance assumptions look reasonable here. The variability of the residuals increases for larger values of $\widehat{y}$, so the constant variance assumption does not look reasonable. And there is a clear pattern to the residuals, rather than a random scatter around 0: as $\widehat{y}$ increases, the residuals are all above 0, then all below 0, then above 0 again.

---

## Concept check

.pull-left[
```{r, echo = F, message=F, fig.align='center', fig.width=6, fig.height=6}
sim_dat <- data.frame(x = seq(-50, 50),
           y = 0.02*seq(-50, 50)^2 + rnorm(101, sd=10))
sim_reg <- lm(y ~ x, data = sim_dat)

data.frame(pred = sim_reg$fitted.values,
           res = sim_reg$residuals) %>%
  ggplot(aes(x = pred, y = res)) +
  geom_point(size=1.5) +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(title = "Residual plot",
       x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$")) +
  theme_bw() +
  theme(text = element_text(size=20))
```
]

.pull-right[
Which assumption looks reasonable?

.abox[
Shape
]

.bbox[
Constant variance
]

.cbox[
Both shape and constant variance
]

.dbox[
Neither shape nor constant variance
]

]


---

## Concept check

.pull-left[
```{r, echo = F, message=F, fig.align='center', fig.width=6, fig.height=6}
sim_dat <- data.frame(x = seq(-50, 50),
           y = 0.02*seq(-50, 50)^2 + rnorm(101, sd=10))
sim_reg <- lm(y ~ x, data = sim_dat)

data.frame(pred = sim_reg$fitted.values,
           res = sim_reg$residuals) %>%
  ggplot(aes(x = pred, y = res)) +
  geom_point(size=1.5) +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(title = "Residual plot",
       x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$")) +
  theme_bw() +
  theme(text = element_text(size=20))
```
]

.pull-right[
Which assumption looks reasonable?

.abox[
Shape
]

.bbox[
Constant variance
]

.cbox[
Both shape and constant variance
]

.dbox[
Neither shape nor constant variance
]

]

Solution: the constant variance assumption looks reasonable, but the shape assumption does not. The variability of the residuals looks pretty constant for different values of $\widehat{y}$. However, there is a clear pattern to the residuals, rather than a random scatter around 0: as $\widehat{y}$ increases, the residuals are generally above 0, then below 0, then above 0 again.

---

## Concept check

.pull-left[
```{r, echo = F, message=F, fig.align='center', fig.width=6, fig.height=6}
data.frame(x = 1:100,
           y = rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(title = "Residual plot",
       x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$")) +
  theme_bw() +
  theme(text = element_text(size=20))
```
]

.pull-right[
Which assumption looks reasonable?

.abox[
Shape
]

.bbox[
Constant variance
]

.cbox[
Both shape and constant variance
]

.dbox[
Neither shape nor constant variance
]

]

---

## Concept check

.pull-left[
```{r, echo = F, message=F, fig.align='center', fig.width=6, fig.height=6}
data.frame(x = 1:100,
           y = rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(title = "Residual plot",
       x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$")) +
  theme_bw() +
  theme(text = element_text(size=20))
```
]

.pull-right[
Which assumption looks reasonable?

.abox[
Shape
]

.bbox[
Constant variance
]

.cbox[
Both shape and constant variance
]

.dbox[
Neither shape nor constant variance
]

]

Solution: Both the shape and constant variance assumptions look reasonable here. This plot is what we hope our residual plots look like: a nice random scatter around 0, with the same variance for different values of $\widehat{y}$.

---

## Formal assumptions for inference

* Shape
* Constant variance
* **Independence**
* Randomness
* Zero mean
* Normality

---

## Independence

**Assumption:** Observations are independent -- one point falling above or below the line has no influence on the location of another point

--

Independence can't be checked with plots. We need to think about how the data were collected.

---

## Independence?

.pull-left[
**Scenario:** You collect a sample of 116 sparrows from Kent island. For each sparrow, you record their wing length and weight. Your data has 116 rows (one for each sparrow), and two columns.
]

.pull-right[
Is it reasonable to assume the observations are independent?

.abox[
Yes
]

.bbox[
No
]
]

---

## Independence?

.pull-left[
**Scenario:** You collect a sample of 116 sparrows from Kent island. For each sparrow, you record their wing length and weight. Your data has 116 rows (one for each sparrow), and two columns.
]

.pull-right[
Is it reasonable to assume the observations are independent?

.abox[
Yes
]

.bbox[
No
]
]

Solution: In this case, it is reasonable to assume the observations are independent. We don't expect that one sparrow's measurements will tell us much about another's. There may be some sparrows in the data which are related, but usually we treat independence as reasonable -- if we only observed pairs of identical twins, then we *would* worry about dependence in the data.

---

## Independence?

.pull-left[
**Scenario:** You go to Kent island, and find a sparrow. For 116 consecutive days, you record its weight and wing length. Your data has 116 rows, and two columns.
]

.pull-right[
Is it reasonable to assume the observations are independent?

.abox[
Yes
]

.bbox[
No
]
]

---

## Independence?

.pull-left[
**Scenario:** You go to Kent island, and find a sparrow. For 116 consecutive days, you record its weight and wing length. Your data has 116 rows, and two columns.
]

.pull-right[
Is it reasonable to assume the observations are independent?

.abox[
Yes
]

.bbox[
No
]
]

Solution: the independence assumption is not reasonable here. A sparrow's measurements on one day will be strongly dependent on what they were the day before.

---

## Class activity

Practice with residual plots and checking assumptions on the Honda Accord data.

[https://sta112-f21.github.io/class_activities/ca_lecture_10.html](https://sta112-f21.github.io/class_activities/ca_lecture_10.html)