---
author: "Dr. Ciaran Evans"
title: Multiple logistic regression
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

## Agenda

* Course logistics:
    * Lab 7 due Monday
    * EC 2 due November 29
    * Project 3 released this weekend
    * Monday: optional work day (EC 2 or Project 3)
    * No more quizzes or assignments (time to focus on Project 3 instead)
* Today: continuing with multiple logistic regression

---

## EDA with the Titanic data

**Titanic data:** 1313 passengers on the Titanic, with the following variables:

* `Survived` (1 = yes, 0 = no)
* `Age` (in years)
* `Sex` (female or male)
* `PClass` (passenger class; 1st, 2nd, or 3rd)

**Question:** After accounting for other variables, is there a relationship between age and survival?

---

## EDA: age and survival

.question[
I want to explore the relationship between age and survival. What plot can I make?
]

---

## EDA: age and survival

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
library(tidyverse)
library(Stat2Data)
data("Titanic")

num_bins <- 8

logodds_table <- Titanic %>%
  drop_na() %>%
  mutate(obs = Survived,
         bin = cut(Age, 
                   breaks = quantile(Age, 
                                     seq(0, 1,
                                         length.out=(num_bins+1))),
                   labels = F, include.lowest=T)) %>%
  group_by(bin) %>%
  summarize(mean_x = mean(Age),
            prop = mean(c(obs, 0.5)),
            num_obs = n()) %>%
  ungroup() %>%
  mutate(logodds = log(prop/(1 - prop)))

logodds_table %>%
  ggplot(aes(x = mean_x,
             y = logodds)) +
  geom_point(size=2) +
  geom_smooth(se=F, method="lm") +
  theme_bw() +
  labs(x = "Average Age",
       y = "Empirical log-odds",
       color = "",) +
  theme(text = element_text(size=20))

```

* Does the shape assumption seem reasonable here?
* What might explain the observed shape?

---

## EDA: age and survival

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}

num_bins <- 8

logodds_table <- Titanic %>%
  drop_na() %>%
  mutate(obs = Survived,
         bin = cut(Age, 
                   breaks = quantile(Age, 
                                     seq(0, 1,
                                         length.out=(num_bins+1))),
                   labels = F, include.lowest=T)) %>%
  group_by(bin) %>%
  summarize(mean_x = mean(Age),
            prop = mean(c(obs, 0.5)),
            num_obs = n()) %>%
  ungroup() %>%
  mutate(logodds = log(prop/(1 - prop)))

logodds_table %>%
  ggplot(aes(x = mean_x,
             y = logodds)) +
  geom_point(size=2) +
  geom_smooth(se=F, method="lm") +
  theme_bw() +
  labs(x = "Average Age",
       y = "Empirical log-odds",
       color = "",) +
  theme(text = element_text(size=20))

```

* One reason for the unusual shape could be that we need to include other variables
* (This could also explain why the slope looks positive for age > 20)

---

## Class activity, Part I

EDA for the relationship between age and survival, broken down by sex and passenger class.

[https://sta112-f21.github.io/class_activities/ca_lecture_37.html](https://sta112-f21.github.io/class_activities/ca_lecture_37.html)

---

## Class activity, Part I

```{r echo=F, message=F, warning=F, fig.width=8, fig.height=4, fig.align='center'}
num_bins <- 6

logodds_table <- Titanic %>%
  drop_na() %>%
  group_by(Sex, PClass) %>%
  mutate(obs = Survived,
         bin = cut(Age, 
                   breaks = quantile(Age, 
                                     seq(0, 1,
                                         length.out=(num_bins+1))),
                   labels = F, include.lowest=T)) %>%
  group_by(bin, Sex, PClass) %>%
  summarize(mean_x = mean(Age),
            prop = mean(c(obs, 0.5)),
            num_obs = n()) %>%
  ungroup() %>%
  mutate(logodds = log(prop/(1 - prop)))

logodds_table %>%
  ggplot(aes(x = mean_x,
             y = logodds,
             color = as.factor(PClass))) +
  geom_point(size=2) +
  geom_smooth(se=F, method="lm") +
  theme_bw() +
  labs(x = "Average Age",
       y = "Empirical log-odds",
       color = "",) +
  theme(text = element_text(size=20)) +
  facet_wrap(~Sex)

```

.question[
Does the shape assumption seem reasonable for each group?
]

---

## Class activity, Part I

```{r echo=F, message=F, warning=F, fig.width=8, fig.height=4, fig.align='center'}
num_bins <- 6

logodds_table <- Titanic %>%
  drop_na() %>%
  group_by(Sex, PClass) %>%
  mutate(obs = Survived,
         bin = cut(Age, 
                   breaks = quantile(Age, 
                                     seq(0, 1,
                                         length.out=(num_bins+1))),
                   labels = F, include.lowest=T)) %>%
  group_by(bin, Sex, PClass) %>%
  summarize(mean_x = mean(Age),
            prop = mean(c(obs, 0.5)),
            num_obs = n()) %>%
  ungroup() %>%
  mutate(logodds = log(prop/(1 - prop)))

logodds_table %>%
  ggplot(aes(x = mean_x,
             y = logodds,
             color = as.factor(PClass))) +
  geom_point(size=2) +
  geom_smooth(se=F, method="lm") +
  theme_bw() +
  labs(x = "Average Age",
       y = "Empirical log-odds",
       color = "",) +
  theme(text = element_text(size=20)) +
  facet_wrap(~Sex)

```

.question[
Does there appear to be a relationship between age and survival, after accounting for sex and class?
]

---

## Class activity, Part I

```{r echo=F, message=F, warning=F, fig.width=8, fig.height=4, fig.align='center'}
num_bins <- 6

logodds_table <- Titanic %>%
  drop_na() %>%
  group_by(Sex, PClass) %>%
  mutate(obs = Survived,
         bin = cut(Age, 
                   breaks = quantile(Age, 
                                     seq(0, 1,
                                         length.out=(num_bins+1))),
                   labels = F, include.lowest=T)) %>%
  group_by(bin, Sex, PClass) %>%
  summarize(mean_x = mean(Age),
            prop = mean(c(obs, 0.5)),
            num_obs = n()) %>%
  ungroup() %>%
  mutate(logodds = log(prop/(1 - prop)))

logodds_table %>%
  ggplot(aes(x = mean_x,
             y = logodds,
             color = as.factor(PClass))) +
  geom_point(size=2) +
  geom_smooth(se=F, method="lm") +
  theme_bw() +
  labs(x = "Average Age",
       y = "Empirical log-odds",
       color = "",) +
  theme(text = element_text(size=20)) +
  facet_wrap(~Sex)

```

.question[
Which combination of age, class, and sex appears to have the highest chance of survival? The lowest? Why might that be?
]

---

## Class activity, Part I

```{r echo=F, message=F, warning=F, fig.width=8, fig.height=4, fig.align='center'}
num_bins <- 6

logodds_table <- Titanic %>%
  drop_na() %>%
  group_by(Sex, PClass) %>%
  mutate(obs = Survived,
         bin = cut(Age, 
                   breaks = quantile(Age, 
                                     seq(0, 1,
                                         length.out=(num_bins+1))),
                   labels = F, include.lowest=T)) %>%
  group_by(bin, Sex, PClass) %>%
  summarize(mean_x = mean(Age),
            prop = mean(c(obs, 0.5)),
            num_obs = n()) %>%
  ungroup() %>%
  mutate(logodds = log(prop/(1 - prop)))

logodds_table %>%
  ggplot(aes(x = mean_x,
             y = logodds,
             color = as.factor(PClass))) +
  geom_point(size=2) +
  geom_smooth(se=F, method="lm") +
  theme_bw() +
  labs(x = "Average Age",
       y = "Empirical log-odds",
       color = "",) +
  theme(text = element_text(size=20)) +
  facet_wrap(~Sex)

```

.question[
Based on the plot, what model should I choose?
]

---

## Class activity, Part I

```{r echo=F, message=F, warning=F, fig.width=8, fig.height=4, fig.align='center'}
num_bins <- 6

logodds_table <- Titanic %>%
  drop_na() %>%
  group_by(Sex, PClass) %>%
  mutate(obs = Survived,
         bin = cut(Age, 
                   breaks = quantile(Age, 
                                     seq(0, 1,
                                         length.out=(num_bins+1))),
                   labels = F, include.lowest=T)) %>%
  group_by(bin, Sex, PClass) %>%
  summarize(mean_x = mean(Age),
            prop = mean(c(obs, 0.5)),
            num_obs = n()) %>%
  ungroup() %>%
  mutate(logodds = log(prop/(1 - prop)))

logodds_table %>%
  ggplot(aes(x = mean_x,
             y = logodds,
             color = as.factor(PClass))) +
  geom_point(size=2) +
  geom_smooth(se=F, method="lm") +
  theme_bw() +
  labs(x = "Average Age",
       y = "Empirical log-odds",
       color = "",) +
  theme(text = element_text(size=20)) +
  facet_wrap(~Sex)

```

$$
\begin{align}
\log \left( \dfrac{\pi}{1 - \pi} \right) &= \beta_0 + \beta_1 \ \text{IsMale} + \beta_2 \ \text{IsSecondClass} + \beta_3 \ \text{IsThirdClass} +  \beta_4 \ \text{Age} \ + \\
& \hspace{0.5cm} \beta_5 \ \text{IsMale} \cdot \text{IsSecondClass} + \beta_6 \ \text{IsMale} \cdot \text{IsThirdClass} \ + \\
& \hspace{0.5cm} \beta_7 \ \text{IsMale} \cdot \text{Age} + \beta_8 \ \text{IsSecondClass} \cdot \text{Age} + \beta_9 \ \text{IsThirdClass} \cdot \text{Age} \ + \\
& \hspace{0.5cm} \beta_{10} \ \text{IsMale} \cdot \text{IsSecondClass} \cdot \text{Age} + \beta_{11} \ \text{IsMale} \cdot \text{IsThirdClass} \cdot \text{Age}
\end{align}
$$

---

## Class activity, Part II

Fit and interpret the model.

[https://sta112-f21.github.io/class_activities/ca_lecture_37.html](https://sta112-f21.github.io/class_activities/ca_lecture_37.html)

---

## Class activity, Part II

$$
\begin{align}
\log \left( \dfrac{\pi}{1 - \pi} \right) &= \beta_0 + \beta_1 \ \text{IsMale} + \beta_2 \ \text{IsSecondClass} + \beta_3 \ \text{IsThirdClass} +  \beta_4 \ \text{Age} \ + \\
& \hspace{0.5cm} \beta_5 \ \text{IsMale} \cdot \text{IsSecondClass} + \beta_6 \ \text{IsMale} \cdot \text{IsThirdClass} \ + \\
& \hspace{0.5cm} \beta_7 \ \text{IsMale} \cdot \text{Age} + \beta_8 \ \text{IsSecondClass} \cdot \text{Age} + \beta_9 \ \text{IsThirdClass} \cdot \text{Age} \ + \\
& \hspace{0.5cm} \beta_{10} \ \text{IsMale} \cdot \text{IsSecondClass} \cdot \text{Age} + \beta_{11} \ \text{IsMale} \cdot \text{IsThirdClass} \cdot \text{Age}
\end{align}
$$

.question[
What is the estimated odds ratio of survival for a 40 year old woman in first class vs. a 40 year old man in first class?
]

--

**Odds ratio:** After simplifying, $\dfrac{1}{\exp\{ \beta_1 + \beta_7 \cdot 40\}}$

So our estimate is $\dfrac{1}{\exp\{ \widehat{\beta_1} + \widehat{\beta_7} \cdot 40\}} = \dfrac{1}{\exp\{ -1.005 - 0.0664 \cdot 40 \}} = 38.9$

---

## Class activity, Part II

**Odds ratio:** After simplifying, $\dfrac{1}{\exp\{ \beta_1 + \beta_7 \cdot 40\}}$

So our estimate is $\dfrac{1}{\exp\{ \widehat{\beta_1} + \widehat{\beta_7} \cdot 40\}} = \dfrac{1}{\exp\{ -1.005 - 0.0664 \cdot 40 \}} = 38.9$

.question[
Does this estimated odds ratio change as age changes?
]

--

**Answer:** Yes; the estimated odds ratio is $\dfrac{1}{\exp\{ -1.005 - 0.0664 \cdot \text{Age} \}}$

As Age increases, so does the estimated odds ratio.

---

## Class activity, Part III

* Is there a relationship between age and survival, after accounting for sex and class?
* After accounting for sex, does the change in the odds of survival associated with a 1-year increase in age depend on class?

[https://sta112-f21.github.io/class_activities/ca_lecture_37.html](https://sta112-f21.github.io/class_activities/ca_lecture_37.html)

---

## Class activity, Part III

Is there a relationship between age and survival, after accounting for sex and class?

.question[
What's our reduced model?
]

--

The reduced model doesn't include age:

$$
\begin{align}
\log \left( \dfrac{\pi}{1 - \pi} \right) &= \beta_0 + \beta_1 \ \text{IsMale} + \beta_2 \ \text{IsSecondClass} + \beta_3 \ \text{IsThirdClass} +\\
& \hspace{0.5cm} \beta_4 \ \text{IsMale} \cdot \text{IsSecondClass} + \beta_5 \ \text{IsMale} \cdot \text{IsThirdClass}
\end{align}
$$

---

## Class activity, Part III

After accounting for sex, does the change in the odds of survival associated with a 1-year increase in age depend on class?

.question[
What's our reduced model?
]

--

The reduced model removes interactions betwen age and class (forcing the slopes to be the same within each sex):

$$
\begin{align}
\log \left( \dfrac{\pi}{1 - \pi} \right) &= \beta_0 + \beta_1 \ \text{IsMale} + \beta_2 \ \text{IsSecondClass} + \beta_3 \ \text{IsThirdClass} +  \beta_4 \ \text{Age} \ + \\
& \hspace{0.5cm} \beta_5 \ \text{IsMale} \cdot \text{IsSecondClass} + \beta_6 \ \text{IsMale} \cdot \text{IsThirdClass} \ + \\
& \hspace{0.5cm} \beta_7 \ \text{IsMale} \cdot \text{Age}
\end{align}
$$