---
author: "Dr. Ciaran Evans"
title: Multicollinearity
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r, include=F}
library(tidyverse)
library(knitr)
library(car)
library(Stat2Data)
data("CountyHealth")

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })
```

## Agenda

* Quiz 4 next Wednesday
    * Covers up through interactions
* MST 121 *or* MST 205 ok for STA 362/363
* Event: interested in getting involved in research?
    * Math faculty give short presentations on their research
    * Manchester 018, Thursday October 21st at 6pm
* Today: multicollinearity (correlated predictors)

---

## CountyHealth data

Can we predict the number of doctors in a county?

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
cty_health <- CountyHealth %>%
  mutate(sqrt_mds = sqrt(MDs))

cty_health %>%
  ggplot(aes(x = Hospitals, y = sqrt_mds)) +
  geom_point(size = 2) +
  labs(x = "Number of hospitals", y = "sqrt(Number of doctors)") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

---

## Using the number of hospitals

```{r, output.lines=10:18}
md_lm <- lm(sqrt_mds ~ Hospitals, data = cty_health)
summary(md_lm)
```

$R^2 = 0.852$, $R^2_{adj} = 0.849$

---

## Using the number of hospital beds

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}

cty_health %>%
  ggplot(aes(x = Beds, y = sqrt_mds)) +
  geom_point(size = 2) +
  labs(x = "Number of hospital beds", y = "sqrt(Number of doctors)") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

---

## Using the number of hospital beds

```{r, output.lines=10:18}
md_lm <- lm(sqrt_mds ~ Beds, data = cty_health)
summary(md_lm)
```

$R^2 = 0.901$, $R^2_{adj} = 0.899$

---

## Using both variables

```{r, output.lines=10:19}
md_lm <- lm(sqrt_mds ~ Hospitals + Beds, data = cty_health)
summary(md_lm)
```

$R^2_{adj} = 0.919$

--

.question[
Adding Hospitals to the model only increased $R^2_{adj}$ from $0.899$ to $0.919$. But we know Hospitals is strongly associated with sqrt(Doctors). Why the small change?
]

---

### Multicollinearity (correlated predictors)

.question[
**Multicollinearity:** one or more predictors is strongly correlated with some combination of the other predictors.
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
cty_health %>%
  ggplot(aes(x = Beds, y = Hospitals)) +
  geom_point(size = 2) +
  labs(x = "Number of hospital beds", y = "Number of hospitals") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

$r = 0.91$

---

### Measuring multicollinearity

**Motivation:** high multicollinearity means at least one predictor can be estimated well by a combination of other predictors

**Variance inflation factor:** Let $X_i$ be a predictor in the model, and $R^2_i$ the coefficient of determination for a model predicting $X_i$ with the other predictors. The **variance inflation factor** for $X_i$ is

$VIF_i = \dfrac{1}{1 - R^2_i}$

More multicollinearity means a higher VIF.

**Rule of thumb:** worried when VIF $> 5$

---

## Example

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
cty_health %>%
  ggplot(aes(x = Beds, y = Hospitals)) +
  geom_point(size = 2) +
  labs(x = "Number of hospital beds", y = "Number of hospitals") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

---

## Example

```{r, output.lines=10:18}
hospital_lm <- lm(Hospitals ~ Beds, data = cty_health)
summary(hospital_lm)
```

$R^2 = 0.827$

VIF for Hospitals $= \dfrac{1}{1 - 0.827} = 5.78$

---

## In R

```{r}
library(car)
md_lm <- lm(sqrt_mds ~ Hospitals + Beds, data = cty_health)
vif(md_lm)
```

---

## Why is multicollinearity a problem?

* Inflates variability of estimated coefficients
* Interpretation of individual terms is difficult

.center[
$\widehat{\sqrt{Doctors}} = 3.58 + 2.58 \ \text{Hospitals} \ + 0.012 \ \text{Beds}$
]

**Usual interpretation:** Holding Beds fixed, an increase of 1 in the number of hospitals is associated with an increase of 2.58 units in $\sqrt{Doctors}$.

**Problem:** When Beds and Hospitals are highly correlated, doesn't make sense to fix one and change the other.

---

## Why is multicollinearity a problem?

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
cty_health %>%
  ggplot(aes(x = Beds, y = Hospitals)) +
  geom_point(size = 2) +
  labs(x = "Number of hospital beds", y = "Number of hospitals") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

.question[
Can't really think about fixing the number of beds and changing the number of hospitals.
]

---

## Class activity

---

## Class activity results

* $\widehat{\text{Hgt97}} = 42.75 + 1.09 \ \text{Hgt96} \hspace{2cm} R^2_{adj} = 0.934$

---

## Class activity results

* $\widehat{\text{Hgt97}} = 42.75 + 1.09 \ \text{Hgt96} \hspace{2cm} R^2_{adj} = 0.934$
* $\widehat{\text{Hgt97}} = 41.49 - 0.88 \ \text{Diam96} + 1.11 \ \text{Hgt96} \hspace{2cm} R^2_{adj} =$

---

## Class activity results

* $\widehat{\text{Hgt97}} = 42.75 + 1.09 \ \text{Hgt96} \hspace{2cm} R^2_{adj} = 0.934$
* $\widehat{\text{Hgt97}} = 41.49 - 0.88 \ \text{Diam96} + 1.11 \ \text{Hgt96} \hspace{2cm} R^2_{adj} = 0.934$

---

## Class activity results

* $\widehat{\text{Hgt97}} = 42.75 + 1.09 \ \text{Hgt96} \hspace{2cm} R^2_{adj} = 0.934$
* $\widehat{\text{Hgt97}} = 41.49 - 0.88 \ \text{Diam96} + 1.11 \ \text{Hgt96} \hspace{2cm} R^2_{adj} = 0.934$
* VIF for Diam96 (and Hgt96) $=$

---

## Class activity results

* $\widehat{\text{Hgt97}} = 42.75 + 1.09 \ \text{Hgt96} \hspace{2cm} R^2_{adj} = 0.934$
* $\widehat{\text{Hgt97}} = 41.49 - 0.88 \ \text{Diam96} + 1.11 \ \text{Hgt96} \hspace{2cm} R^2_{adj} = 0.934$
* VIF for Diam96 (and Hgt96) $= 5.06$
* **Issue with interpretation:** 

---

## Class activity results

* $\widehat{\text{Hgt97}} = 42.75 + 1.09 \ \text{Hgt96} \hspace{2cm} R^2_{adj} = 0.934$
* $\widehat{\text{Hgt97}} = 41.49 - 0.88 \ \text{Diam96} + 1.11 \ \text{Hgt96} \hspace{2cm} R^2_{adj} = 0.934$
* VIF for Diam96 (and Hgt96) $= 5.06$
* **Issue with interpretation:** doesn't really make sense to fix Hgt96 and change Diam96