---
author: "Dr. Ciaran Evans"
title: Regression with a categorical predictor
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

## Agenda

* Math and Stats department town hall **today**, 4:00--4:45pm in Manchester 016. Pizza included!
* Quiz 3 next Wednesday (covers up through today)
* Solutions for lab 3 and 4 on Canvas
* Today: regression with a categorical predictor
    
---

## Simple linear regression

**Model:**

.center[
$y = \beta_0 + \beta_1 x + \varepsilon$
]

where $x$ and $y$ are both **quantitative**.

$\beta_0 + \beta_1 x$ is the mean value of $y$, given $x$.

.question[
What if $x$ is *categorical*?
]

---

## Categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
library(palmerpenguins)
library(tidyverse)

penguins_no_nas <- penguins %>%
  drop_na()

penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot() +
  theme_classic() +
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 20))
```

* How do we model the relationship between species and bill length?
* Can we test for a relationship between bill length and species?

---

## Categorical predictor

Imagine population like this:

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0.5) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38.8, xend = 1.4, yend = 38.8), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 48.8, xend = 2.4, yend = 48.8), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47.6, xend = 3.4, yend = 47.6), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 20))
```

If $\text{species} = \text{Adelie}, \hspace{1cm} \text{bill length} = \text{mean Adelie bill length} + \varepsilon$

If $\text{species} = \text{Chinstrap}, \hspace{0.3cm} \text{bill length} = \text{mean Chinstrap bill length} + \varepsilon$

If $\text{species} = \text{Gentoo}, \hspace{1cm} \text{bill length} = \text{mean Gentoo bill length} + \varepsilon$

---

## Categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=5, fig.height=3}

library(latex2exp)

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38, xend = 1.4, yend = 38), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 50, xend = 2.4, yend = 50), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47, xend = 3.4, yend = 47), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 20)) +
  theme(axis.title.y = element_blank()) +
  scale_y_continuous(breaks = c(38),
                     labels = c(TeX("$\\beta_0")))
```

If $\text{species} = \text{Adelie}, \hspace{1cm} \text{bill length} = \beta_0 + \varepsilon$

* $\beta_0$ = true mean bill length of Adelie penguins (in the population)

If $\text{species} = \text{Chinstrap}, \hspace{0.3cm} \text{bill length} = \text{mean Chinstrap bill length} + \varepsilon$

If $\text{species} = \text{Gentoo}, \hspace{1cm} \text{bill length} = \text{mean Gentoo bill length} + \varepsilon$

---

## Categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=5, fig.height=3}

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38, xend = 1.4, yend = 38), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 50, xend = 2.4, yend = 50), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47, xend = 3.4, yend = 47), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 20)) +
  theme(axis.title.y = element_blank()) +
  scale_y_continuous(breaks = c(38, 50),
                     labels = c(TeX("$\\beta_0"), 
                                TeX("$\\beta_0 + \\beta_1"))) +
  geom_segment(aes(x = 1.45, y = 38,
                   xend = 1.45, yend = 50),
               arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x = 1.3, y = 44, 
           label = TeX("$\\beta_1", output='character'), 
           parse=TRUE, size = 7)
```

If $\text{species} = \text{Adelie}, \hspace{1cm} \text{bill length} = \beta_0 + \varepsilon$

* $\beta_0$ = true mean bill length of Adelie penguins (in the population)

If $\text{species} = \text{Chinstrap}, \hspace{0.3cm} \text{bill length} = \beta_0 + \beta_1 + \varepsilon$

* $\beta_0 + \beta_1$ = true mean bill length of Chinstrap penguins (in the population)

If $\text{species} = \text{Gentoo}, \hspace{1cm} \text{bill length} = \text{mean Gentoo bill length} + \varepsilon$

---

## Categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=5, fig.height=3}

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38, xend = 1.4, yend = 38), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 50, xend = 2.4, yend = 50), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47, xend = 3.4, yend = 47), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 20)) +
  theme(axis.title.y = element_blank()) +
  scale_y_continuous(breaks = c(38, 50, 47),
                     labels = c(TeX("$\\beta_0"), 
                                TeX("$\\beta_0 + \\beta_1"), 
                                TeX("$\\beta_0 + \\beta_2"))) +
  geom_segment(aes(x = 1.45, y = 38,
                   xend = 1.45, yend = 50),
               arrow = arrow(length = unit(0.3, "cm"))) +
  geom_segment(aes(x = 2.45, y = 38,
                   xend = 2.45, yend = 47),
               arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x = 1.3, y = 44, 
           label = TeX("$\\beta_1", output='character'), 
           parse=TRUE, size = 7) +
  annotate("text", x = 2.3, y = 43, 
           label = TeX("$\\beta_2", output='character'), 
           parse=TRUE, size = 7)
```

If $\text{species} = \text{Adelie}, \hspace{1cm} \text{bill length} = \beta_0 + \varepsilon$

* $\beta_0$ = true mean bill length of Adelie penguins (in the population)

If $\text{species} = \text{Chinstrap}, \hspace{0.3cm} \text{bill length} = \beta_0 + \beta_1 + \varepsilon$

* $\beta_0 + \beta_1$ = true mean bill length of Chinstrap penguins (in the population)

If $\text{species} = \text{Gentoo}, \hspace{1cm} \text{bill length} = \beta_0 + \beta_2 + \varepsilon$

* $\beta_0 + \beta_2$ = true mean bill length of Gentoo penguins (in the population)

---

## Categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38, xend = 1.4, yend = 38), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 50, xend = 2.4, yend = 50), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47, xend = 3.4, yend = 47), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 20)) +
  theme(axis.title.y = element_blank()) +
  scale_y_continuous(breaks = c(38, 50, 47),
                     labels = c(TeX("$\\beta_0"), 
                                TeX("$\\beta_0 + \\beta_1"), 
                                TeX("$\\beta_0 + \\beta_2"))) +
  geom_segment(aes(x = 1.45, y = 38,
                   xend = 1.45, yend = 50),
               arrow = arrow(length = unit(0.3, "cm"))) +
  geom_segment(aes(x = 2.45, y = 38,
                   xend = 2.45, yend = 47),
               arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x = 1.3, y = 44, 
           label = TeX("$\\beta_1", output='character'), 
           parse=TRUE, size = 7) +
  annotate("text", x = 2.3, y = 43, 
           label = TeX("$\\beta_2", output='character'), 
           parse=TRUE, size = 7)
```

Putting it together:

.center[
$\text{bill length} = \begin{cases} \beta_0 + \varepsilon & \text{species = Adelie} \\ \beta_0 + \beta_1 + \varepsilon & \text{species = Chinstrap} \\ \beta_0 + \beta_2 + \varepsilon & \text{species = Gentoo} \end{cases}$
]

--

.question[
Can we write this more concisely?
]

---

## Indicator variables

Let

.center[
$\text{IsChinstrap} = \begin{cases} 0 & \text{species } \neq \text{Chinstrap} \\ 1 & \text{species = Chinstrap} \end{cases}$
]

.center[
$\text{IsGentoo} = \begin{cases} 0 & \text{species } \neq \text{Gentoo} \\ 1 & \text{species = Gentoo} \end{cases}$
]

Then:

| Species | IsChinstrap | IsGentoo |
| --- | --- | --- |
| Adelie | 0 | 0 |
| Chinstrap | 1 | 0 |
| Gentoo | 0 | 1 |

---

## Indicator variables

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38, xend = 1.4, yend = 38), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 50, xend = 2.4, yend = 50), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47, xend = 3.4, yend = 47), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 20)) +
  theme(axis.title.y = element_blank()) +
  scale_y_continuous(breaks = c(38, 50, 47),
                     labels = c(TeX("$\\beta_0"), 
                                TeX("$\\beta_0 + \\beta_1"), 
                                TeX("$\\beta_0 + \\beta_2"))) +
  geom_segment(aes(x = 1.45, y = 38,
                   xend = 1.45, yend = 50),
               arrow = arrow(length = unit(0.3, "cm"))) +
  geom_segment(aes(x = 2.45, y = 38,
                   xend = 2.45, yend = 47),
               arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x = 1.3, y = 44, 
           label = TeX("$\\beta_1", output='character'), 
           parse=TRUE, size = 7) +
  annotate("text", x = 2.3, y = 43, 
           label = TeX("$\\beta_2", output='character'), 
           parse=TRUE, size = 7)
```

.center[
$\text{bill length} = \begin{cases} \beta_0 + \varepsilon & \text{species = Adelie} \\ \beta_0 + \beta_1 + \varepsilon & \text{species = Chinstrap} \\ \beta_0 + \beta_2 + \varepsilon & \text{species = Gentoo} \end{cases}$
]

Or, more concisely:

.center[
$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$
]

---

## Estimated model

.center[
$\widehat{\text{bill length}} = \widehat{\beta}_0 + \widehat{\beta}_1 \text{IsChinstrap} + \widehat{\beta}_2 \text{IsGentoo}$

$\hspace{4cm} = 38.82 + 10.01 \ \text{IsChinstrap} + 8.74 \ \text{IsGentoo}$
]

* What is the estimated average bill length for each of the three species of penguin?
* There are 333 penguins in the dataset used to fit the model. How many degrees of freedom does the fitted model have?

---

## Estimated model

.center[
$\widehat{\text{bill length}} = \widehat{\beta}_0 + \widehat{\beta}_1 \text{IsChinstrap} + \widehat{\beta}_2 \text{IsGentoo}$

$\hspace{4cm} = 38.82 + 10.01 \ \text{IsChinstrap} + 8.74 \ \text{IsGentoo}$
]

| Species | $\widehat{\text{bill length}}$ |
| --- | --- |
| Adelie | 38.82 | 
| Chinstrap | 38.82 + 10.01 = 48.83 | 
| Gentoo | 38.82 + 8.74 = 47.56 |

--

**Degrees of freedom:** 333 - 3 = 330
* (Lose one degree of freedom for each parameter estimated)

---

## Fitting the model in R

R will automatically create indicator variables:

```{r}
lm(bill_length_mm ~ species, data = penguins_no_nas)
```

---

## Model assumptions

.center[
$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$
]

where $\varepsilon \sim N(0, \sigma_{\varepsilon}^2)$ and the $\varepsilon$ are independent of each other.

* No notion of shape assumption with single categorical predictor
* Still want constant variance and normality assumptions

---

## Assessing constant variance

.pull-left[
```{r echo=F, fig.align='center', fig.width=6, fig.height=4}
penguin_lm <- lm(bill_length_mm ~ species, 
                 data = penguins_no_nas)

penguins_no_nas %>%
  mutate(resids = residuals(penguin_lm)) %>%
  ggplot(aes(x = species, y = resids)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = "Species", 
       y = "Residuals") +
  theme(text = element_text(size = 20))
```
]

.pull-right[
```{r echo=F, message=F, warning=F}
library(knitr)
penguins_no_nas %>%
  mutate(resids = residuals(penguin_lm)) %>%
  group_by(species) %>%
  summarize(residual_std_dev = round(sd(resids), 2)) %>%
  kable()
```
]

Looking for similar variability between groups. Rule of thumb: not too concerned about constant variance if

.center[
$\dfrac{\text{largest standard deviation}}{\text{smallest standard deviation}} < 2$
]

---

## Assessing normality

Can use a QQ plot again:

```{r echo=F, fig.align='center', fig.width=6, fig.height=4}
penguins_no_nas %>%
  mutate(resids = residuals(penguin_lm)) %>%
  ggplot(aes(sample = resids)) +
  geom_qq() +
  geom_qq_line() +
  theme_bw() +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed residual quantiles") +
  theme(text = element_text(size = 20))
```

---

## Class activity

Practice regression with a single categorical predictor, with the `drosophila` data:

[https://sta112-f21.github.io/class_activities/ca_lecture_16.html](https://sta112-f21.github.io/class_activities/ca_lecture_16.html)