---
author: "Dr. Ciaran Evans"
title: Intervals for prediction
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

## Agenda

* HW 2 grades released; look at comments, contact Johe or me with any questions
* Today: 
    * intervals for prediction
    * Quiz 2
* Friday: wrapping up simple linear regression
* Next week:
    * beginning multiple linear regression
    * Project 1 released
---

## Recap: confidence interval for $\widehat{\beta}_1$

.center[
$\widehat{\beta}_1 \pm t^* SE_{\widehat{\beta}_1}$
]

* $t^*$ = critical value of $t_{n-2}$ distribution
* $SE_{\widehat{\beta}_1}$ = estimated standard error of $\widehat{\beta}_1$

---

## Confidence intervals for the response

.center[
$\text{price} = \beta_0 + \beta_1 \text{mileage} + \varepsilon$
]

<ins>Question 1:</ins> What is a confidence interval for the average price of used Honda Accords with 50000 miles?

<br>

<ins>Question 2:</ins> I have a used Honda Accord with 50000 miles. What is a confidence interval for the price I can sell it for?

---

## Confidence intervals for the response

.center[
$\text{price} = \beta_0 + \beta_1 \text{mileage} + \varepsilon$
]

<ins>Question 1:</ins> What is a confidence interval for the average price of used Honda Accords with 50000 miles?
* **Confidence interval for mean response**

<br>

<ins>Question 2:</ins> I have a used Honda Accord with 50000 miles. What is a confidence interval for the price I can sell it for?
* **Confidence interval for individual response** (aka **prediction interval**)

---

## Confidence intervals for the response

.center[
$y = \beta_0 + \beta_1 x + \varepsilon$
]

**Confidence interval for mean response:** When $x = x^*$, the mean response is $\mu_y = \beta_0 + \beta_1 x^*$. What is a confidence interval for $\mu_y$, given $x^*$?

<br>

**Prediction interval:** I observe a new data point with $x = x^*$. What is a confidence interval for $y$?

---

## Confidence intervals for the response

.center[
$y = \beta_0 + \beta_1 x + \varepsilon$

$\widehat{y} = \widehat{\beta}_0 + \widehat{\beta}_1 x$
]

**Confidence interval for mean response:** When $x = x^*$, a confidence interval for $\mu_y = \beta_0 + \beta_1 x^*$ is

.center[
$\widehat{y} \pm t^* SE_{\widehat{\mu}}$
]

<br>

**Prediction interval:** When $x = x^*$, a confidence interal for a new response $y$ is

.center[
$\widehat{y} \pm t^* SE_{\widehat{y}}$
]

---

## Confidence intervals for the response

.center[
$y = \beta_0 + \beta_1 x + \varepsilon$

$\widehat{y} = \widehat{\beta}_0 + \widehat{\beta}_1 x$
]

**Confidence interval for mean response:** $\widehat{y} \pm t^* SE_{\widehat{\mu}}$

<br>

**Prediction interval:** $\widehat{y} \pm t^* SE_{\widehat{y}}$

<br>

$SE_{\widehat{\mu}} = \widehat{\sigma}_{\varepsilon} \sqrt{\dfrac{1}{n} + \dfrac{(x^* - \overline{x})^2}{\sum_i (x_i - \overline{x})^2}}$

$SE_{\widehat{y}} = \widehat{\sigma}_{\varepsilon} \sqrt{1 + \dfrac{1}{n} + \dfrac{(x^* - \overline{x})^2}{\sum_i (x_i - \overline{x})^2}}$

---

## Confidence intervals for the response

.center[
$y = \beta_0 + \beta_1 x + \varepsilon$

$\widehat{y} = \widehat{\beta}_0 + \widehat{\beta}_1 x$
]

**Confidence interval for mean response:** $\widehat{y} \pm t^* SE_{\widehat{\mu}}$

**Prediction interval:** $\widehat{y} \pm t^* SE_{\widehat{y}}$

$SE_{\widehat{\mu}} = \widehat{\sigma}_{\varepsilon} \sqrt{\dfrac{1}{n} + \dfrac{(x^* - \overline{x})^2}{\sum_i (x_i - \overline{x})^2}}$

$SE_{\widehat{y}} = \widehat{\sigma}_{\varepsilon} \sqrt{1 + \dfrac{1}{n} + \dfrac{(x^* - \overline{x})^2}{\sum_i (x_i - \overline{x})^2}}$

* $SE_{\widehat{y}} > SE_{\widehat{\mu}}$
* $SE_{\widehat{y}}^2 = SE_{\widehat{\mu}}^2 + \widehat{\sigma}_{\varepsilon}^2$
* Both depend on how close $x^*$ is to $\overline{x}$

---

## Impact of $x^* - \overline{x}$

Accord data:

```{r echo=F, include=F}
library(Stat2Data)
library(tidyverse)
data("AccordPrice")
```

.pull-left[

```{r echo=F, message=F, fig.align='center', fig.width=6, fig.height=4}
p <- AccordPrice %>%
  ggplot(aes(x = Mileage, y = Price)) +
  geom_point(alpha = 0.5, size=2)

for(i in 1:30){
  accord_samp <- AccordPrice %>%
    sample_n(30, replace=T)
  
  samp_lm <- lm(Price ~ Mileage, data = accord_samp)
  
  p <- p +
    geom_abline(slope = coef(samp_lm)[2],
                intercept = coef(samp_lm)[1],
                alpha = 0.7)
}

p <- p +
  geom_smooth(method = "lm", se=F, lwd=2) +
  theme_bw() + 
  labs(x = "Mileage (in 1000's)",
       y = "Price (in $1000's)") +
  theme(text = element_text(size = 20))

p
```

Blue line = estimated line from observed data
Black lines = estimated lines from other possible samples
]

.pull-right[
```{r echo=F, message=F, fig.align='center', fig.width=6, fig.height=4}
AccordPrice %>%
  ggplot(aes(x = Mileage, y = Price)) +
  geom_point(size=2) +
  geom_smooth(method="lm", se=T) +
  theme_bw() +
  labs(x = "Mileage (in 1000's)",
       y = "Price (in $1000's)") +
  theme(text = element_text(size = 20))
```

With `geom_line(method="lm", se=TRUE)` (displays confidence interval for mean response at each point)
]

* There is more variability in the fitted line for larger $x - \overline{x}$
* For similar reasons, points with $x - \overline{x}$ have more potential to be influential

---

## Accord data

Interested in used Accords with 50000 miles:
.center[
$\widehat{\text{price}} = 20.8 - 0.12(50) =$ `$14,800`
]

**95% Confidence interval for mean response:**

```{r}
accord_lm <- lm(Price ~ Mileage, data = AccordPrice)
predict(accord_lm, 
        newdata = data.frame(Mileage=50),
        interval = "confidence",
        level = 0.95)
```

**Interpretation:** We are 95% confident that the true mean price of used Honda Accords with 50000 miles is between `$13,658` and `$15,980`.

---

## Accord data

Interested in used Accords with 50000 miles:
.center[
$\widehat{\text{price}} = 20.8 - 0.12(50) =$ `$14,800`
]

**95% Prediction interval:**

```{r}
accord_lm <- lm(Price ~ Mileage, data = AccordPrice)
predict(accord_lm, 
        newdata = data.frame(Mileage=50),
        interval = "prediction",
        level = 0.95)
```

**Interpretation:** We are 95% confident that the sales price of a used Honda Accord with 50000 miles will be between `$8,394` and `$21,244`.

---

## Class activity

[https://sta112-f21.github.io/class_activities/ca_lecture_14.html](https://sta112-f21.github.io/class_activities/ca_lecture_14.html)