---
---
---
title: "Lab 6"
author: "Your Name"
date: "10/18/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
team_data <- read_csv("https://sta112-f21.github.io/labs/team_data.csv")
```

## Exercise 1

```{r}

```

Type your answer here...

## Exercise 2

Type your answer here...

## Exercise 3


Type your answer here...

## Exercise 4

```{r}

```

Type your answer here...

## Exercise 5

```{r}

```

Type your answer here...

## Exercise 6

```{r}

```

Type your answer here...

## Exercise 7


Type your answer here...

## Exercise 8


Type your answer here...

## Exercise 9

```{r}

```

Type your answer here...

## Exercise 10


Type your answer here...

## Exercise 11

```{r}

```

Type your answer here...

## Exercise 12

```{r}

```

Type your answer here...

## Exercise 13

```{r}

```

Type your answer here...

## Exercise 14

```{r}

```

Type your answer here...

## Exercise 15


Type your answer here...


## Exercise 16

```{r}

```

Type your answer here...

## Exercise 17

```{r}

```

Type your answer here...
