---
---
---
title: "Lab 3"
author: "Your Name"
date: "9/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(palmerpenguins)
library(tidyverse)
```

## Exercise 1

```{r}

```

Type your answer to question 1 here...

## Exercise 2

```{r}

```

Type your answer to question 2 here...

## Exercise 3

```{r}

```

## Exercise 4

Type your answer to question 4 here...

## Exercise 5

```{r}

```


## Exercise 6

Type your answer to question 6 here...

## Exercise 7

```{r}

```

## Exercise 8

Type your answer to question 8 here...

## Exercise 9

```{r}

```

## Exercise 10

```{r}

```


## Exercise 11

Type your answer to question 11 here...

## Exercise 12

```{r}

```

Type your answer to question 12 here...

## Exercise 13

* *Data matching:* 
* *Exhaustive:* 
* *Skeptical:* 
* *Second-order:* 
* *Transparency:* 
* *Reproducibility:*  

## Citations and acknowledgments

Cite any outside resources used, and acknowledge any collaborators