---
title: "Class Activity, September 29"
output: 
  tufte::tufte_html:
    css: "../labs/lab.css"
    tufte_variant: "envisioned"
    highlight: pygments
link-citations: yes
---

In this activity, you will practice hypothesis tests for a relationship between a categorical predictor and a quantitative response.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
drosophila <- read_csv("https://sta112-f21.github.io/homework/drosophila.csv")
diet_lm <- lm(WingLength ~ Diet, data = drosophila)
```

## Setup

This activity investigates the relationship between diet and wing length for the drosophila in HW 2.

Download the activity template: [activity template](https://sta112-f21.github.io/class_activities/ca_lecture_17_template.Rmd)

We can visualize the relationship between diet and wing length with boxplots:

```{r echo=F, fig.align='center'}
drosophila %>%
  ggplot(aes(x = Diet, y = WingLength)) +
  geom_boxplot() +
  theme_bw() +
  theme(text = element_text(size = 20))
``` 

We will model this relationship as

$\text{WingLength} = \beta_0 + \beta_1 \text{IsRich} + \beta_2 \text{IsStandard} + \varepsilon$,

where $\text{IsRich}$ and $\text{IsStandard}$ are indicator variables for Diet:

```{marginfigure}
There are lots of different notation for indicator variables. For example, R would call these indicators `Dietrich` and `Dietstandard`. Other notation includes $I_{\text{rich}}$ and $I_{\text{standard}}$.
```

$\text{IsRich} = \begin{cases} 0 & \text{Diet } \neq \text{rich} \\ 1 & \text{Diet } = \text{rich} \end{cases}$

$\text{IsStandard} = \begin{cases} 0 & \text{Diet } \neq \text{standard} \\ 1 & \text{Diet } = \text{standard} \end{cases}$


## Questions

1. After fitting the model above, we calculate the sums of squares: $SSModel = 6.09$ and $SSE = 7.62$. There are 1372 flies in the data. Using this information, fill in the ANOVA table:

| Source | df | SS | MS | F |
| --- | --- | --- | --- | --- |
| Model | |  | | |
| Residual | |  |  | |
| Total |  | | | |

2. We can also get the ANOVA table for the fitted model in R. Use the `anova` function to get the ANOVA table and the p-value for the hypotheses $H_0: \beta_1 = \beta_2 = 0$ vs. $H_A: \text{one of } \beta_1, \beta_2 \neq 0$. Check that your ANOVA table from question 1 matches the R output (note that R doesn't include the "Total" row in the table, and adds a column for the p-value).

<SCRIPT>
function ShowAndHide(divid) {
    var x = document.getElementById(divid);
    if (x.style.display == 'none') {
        x.style.display = 'block';
    } else {
        x.style.display = 'none';
    }
}
</SCRIPT>

<BUTTON ONCLICK="ShowAndHide('q2_solution')">Solution</BUTTON>

<DIV ID="q2_solution" STYLE="display:none">
```r
diet_lm <- lm(WingLength ~ Diet, data = drosophila)
anova(diet_lm)
```
</DIV>

3. Based on the p-value in question 2 and the boxplots above, do you think there is a relationship between diet and wing length for fruit flies?

4. On Monday, we checked the constant variance and normality assumptions for this model. Do the diagnostic plots from Monday change your conclusions for question 3?