---
title: "Class Activity, November 29"
output: 
  tufte::tufte_html:
    css: "../labs/lab.css"
    tufte_variant: "envisioned"
    highlight: pygments
link-citations: yes
---

```{r, include=F}
library(tidyverse)
library(Stat2Data)
data("Titanic")
```

## Setup

In this activity, we will work with the `Titanic` data.

Download the activity template: [activity template](https://sta112-f21.github.io/class_activities/ca_lecture_38_template.Rmd)


## Questions

From exploratory data analysis, we modeled the relationship between age, sex, class, and survival as 

$$
\begin{align}
(1) \hspace{0.5cm} \log \left( \dfrac{\pi}{1 - \pi} \right) &= \beta_0 + \beta_1 \ \text{IsMale} + \beta_2 \ \text{IsSecondClass} + \beta_3 \ \text{IsThirdClass} +  \beta_4 \ \text{Age} \ + \\
& \hspace{0.5cm} \beta_5 \ \text{IsMale} \cdot \text{IsSecondClass} + \beta_6 \ \text{IsMale} \cdot \text{IsThirdClass} \ + \\
& \hspace{0.5cm} \beta_7 \ \text{IsMale} \cdot \text{Age} + \beta_8 \ \text{IsSecondClass} \cdot \text{Age} + \beta_9 \ \text{IsThirdClass} \cdot \text{Age} \ + \\
& \hspace{0.5cm} \beta_{10} \ \text{IsMale} \cdot \text{IsSecondClass} \cdot \text{Age} + \beta_{11} \ \text{IsMale} \cdot \text{IsThirdClass} \cdot \text{Age}
\end{align}
$$

We want to use that model to answer two questions:

**First question:** Is there a relationship between age and survival, after accounting for sex and class?

**Second question:** After accounting for sex, does the change in the odds of survival associated with a 1-year increase in age depend on class?

We will use a likelihood ratio test to investigate these two questions. Our reduced model for the first question is

$$
\begin{align}
(2) \hspace{0.5cm} \log \left( \dfrac{\pi}{1 - \pi} \right) &= \beta_0 + \beta_1 \ \text{IsMale} + \beta_2 \ \text{IsSecondClass} + \beta_3 \ \text{IsThirdClass} +\\
& \hspace{0.5cm} \beta_4 \ \text{IsMale} \cdot \text{IsSecondClass} + \beta_5 \ \text{IsMale} \cdot \text{IsThirdClass}
\end{align}
$$

and our reduced model for the second question is

$$
\begin{align}
(3) \hspace{0.5cm} \log \left( \dfrac{\pi}{1 - \pi} \right) &= \beta_0 + \beta_1 \ \text{IsMale} + \beta_2 \ \text{IsSecondClass} + \beta_3 \ \text{IsThirdClass} +  \beta_4 \ \text{Age} \ + \\
& \hspace{0.5cm} \beta_5 \ \text{IsMale} \cdot \text{IsSecondClass} + \beta_6 \ \text{IsMale} \cdot \text{IsThirdClass} \ + \\
& \hspace{0.5cm} \beta_7 \ \text{IsMale} \cdot \text{Age}
\end{align}
$$

1. Fit the full model (Equation (1) above), and calculate the deviance (what R calls "residual deviance").

<SCRIPT>
function ShowAndHide(divid) {
    var x = document.getElementById(divid);
    if (x.style.display == 'none') {
        x.style.display = 'block';
    } else {
        x.style.display = 'none';
    }
}
</SCRIPT>

<BUTTON ONCLICK="ShowAndHide('q1_solution')">Solution</BUTTON>

<DIV ID="q1_solution" STYLE="display:none">

```{r}
titanic_no_nas <- Titanic %>% 
  drop_na()

full_glm <- glm(Survived ~ Sex*PClass*Age, 
                data = titanic_no_nas,
                family = binomial)
summary(full_glm)
```

The deviance for the full model is 639.64

</DIV>

2. Fit the reduced model for the first question (Equation (2)), and calculate the test statistic: $G$ = deviance for reduced model - deviance for full model.

<BUTTON ONCLICK="ShowAndHide('q2_solution')">Solution</BUTTON>

<DIV ID="q2_solution" STYLE="display:none">

```{r}
reduced_glm_1 <- glm(Survived ~ Sex*PClass, 
                data = titanic_no_nas,
                family = binomial)

summary(reduced_glm_1)
```

The deviance for the reduced model is 697.84, so the test statistic is $G = 697.84 - 639.64 = 58.2$

</DIV>

3. We know that $G$ follows a $\chi^2$ distribution; what are the degrees of freedom?

<BUTTON ONCLICK="ShowAndHide('q3_solution')">Solution</BUTTON>

<DIV ID="q3_solution" STYLE="display:none">

The degrees of freedom is the number of parameters tested. In this case, we are testing 6 parameters, so we have 6 degrees of freedom

</DIV>

4. Fill in the following code to calculate a p-value for the first question:

```r
pchisq(..., df = ..., lower.tail=F)
```

<BUTTON ONCLICK="ShowAndHide('q4_solution')">Solution</BUTTON>

<DIV ID="q4_solution" STYLE="display:none">

```{r}
pchisq(58.2, df = 6, lower.tail=F)
```

</DIV>

5. What do you conclude about the first question?

<BUTTON ONCLICK="ShowAndHide('q5_solution')">Solution</BUTTON>

<DIV ID="q5_solution" STYLE="display:none">

The p-value is very close to 0, so we have strong evidence that there is a relationship between age and survival, after accounting for sex and class.

</DIV>

6. Carry out a likelihood ratio test to address the second question, and report your results.